### 2023-03-11 outputs/paper/paper.qmd
Init

---
title: "The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?"
format: pdf
author: "Sebastian Rodriguez^[University of Toronto, Faculty of Information]"
abstract: "The Toronto Paramedic Services provides all pre-hospital care to the city’s inhabitants. Recent criticism of the service called into question its performance, highlighting staff shortages and institutional challenges that the city's paramedics face. This paper investigates the Toronto Paramedic Services’ efficacy by comparing its emergency response times to the standards set by provincial and municipal governments. It reveals that Toronto's paramedics meet and often exceed response time standards, and provide stroke and STEMI heart attack patients with a chance of survival greater than the provincial average." 
date: today
date-format: long
editor: visual
toc: TRUE
number-sections: TRUE
bibliography: references.bib
thanks: "Code and data supporting this analysis is available at: https://github.com/seb646/toronto-paramedic-responses"
---

```{r}
#| messages: false
#| echo: false
#| warning: false
#| output: false
#### Preamble ####
# Title: The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses

#### Workspace set-up ####
# Define required libraries 
library(lubridate)
library(tidyverse)
library(dplyr)
library(knitr)
library(janitor)
library(scales)
library(RColorBrewer)
library(ggplot2)
library(kableExtra)
library(here)
library(readr)

# Fetch the the clean response data
clean_response_data <- readr::read_csv(here::here("inputs/data/clean_response_data.csv"), show_col_types = FALSE)

# Fetch clean performance data
clean_performance_data <- readr::read_csv(here::here("inputs/data/clean_performance_data.csv"), show_col_types = FALSE)

# Fetch the clean incident response
clean_incident_data <- readr::read_csv(here::here("inputs/data/clean_incident_data.csv"), show_col_types = FALSE)
```

## Introduction

A city's emergency medical services (EMS) are integral to its health care system. The paramedics and emergency medical technicians working for the service provide pre-hospital assessment, stabilisation, and care to people needing medical assistance. These professionals respond to every call from the public for aid, from broken bones and heart attacks to car accidents and house fires. Current emergency medical services operate on the principle of the "golden hour," which is the 60 minutes between an initial injury and arrival at a trauma center when a patient has the highest statistical chance of survival [@cite_golden_hour]. This "golden hour" is the foundation of all emergency medical responses and requires medical personnel to arrive on scene in under 10 minutes [@cite_golden_hour_10_mins].

In Toronto, inhabitants receive emergency medical assistance through the Toronto Paramedic Services. As of 2020, the Toronto Paramedic Services employed 1,186 paramedics operating 234 ambulances out of 45 stations across the city [@cite_toronto_paramedic_2020_report]. The service provides all land ambulatory responses within the city, including on-scene treatment and stabilisation of patients, transport to a hospital, and transport between hospitals. The Province of Ontario and the City of Toronto set specific requirements for the length of time an ambulance should take when responding to an emergency call, which can be anywhere between 6 to 25 minutes. While the service strives to achieve these mandated goals, they face staffing, budget, and other institutional challenges that impact its ability to provide care. Recently, the service gained attention for slower response times because of staffing shortages in hospitals and in the service itself [@cite_paramedic_shortages]. These staffing shortages decrease the number of available paramedics while simultaneously increasing the time it takes for a patient to enter a hospital, tying up paramedics at hospitals. The recent COVID-19 pandemic also impacted the service, with most paramedics stating that morale is "either worse or as bad as it has ever been" [@cite_paramedic_morale].

Despite the challenges that Toronto's paramedics face, they must respond to emergencies quickly to provide patients with the best chances of survival. This paper investigates the response times of the Toronto Paramedic Service compared to the mandated requirements set by the local and provincial governments. Specifically, the paper aims to place response times within a larger context of the service's responsibility and interrogate the relevance of the "golden hour" principle to determine the efficacy of the service.

The findings indicated a growing trend in the number of emergency medical calls, an increase of approximately 10,000 per year since 2014 [@cite_emergency_care_data]. Despite this increase, the Toronto Paramedic Services met and often exceeded the response times set by both the Province of Ontario and the City of Toronto. The paper also revealed a significant lack of paramedic metrics from the data sets available on Open Data Toronto. Without metrics on case-by-case response times and patient outcomes, other sources were required to inform data analysis. Overall, the response time of Toronto's paramedics contributed to stroke and mortality rates lower than Ontario's average, giving the city's inhabitants greater chances of survival when suffering from life-threatening injuries. \newpage

## Paramedic Response Data

### Scope of Responsibility

Before investigating the response times and efficacy of the Toronto Paramedic Services, it is important to understand the service's breadth of responsibility. With a population of nearly 3 million people, Toronto requires a medical response to roughly 300,000 emergency calls every year. As the city's population grows, so too does the need for medical assistance. There was a steady increase of approximately 10,000 emergency medical requests per year from 2014 to 2019, as outlined by [@fig-1; @cite_emergency_care_data]. This trend ended briefly in 2020 when the COVID-19 pandemic hit Toronto, placing an incredible strain on the city's healthcare system. Due to a concern about COVID-19 infection, many people delayed or avoided medical care throughout the pandemic [@cite_avoiding_hospitals].

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Number of Emergency Medical Responses in Toronto (by year since 2014)"
#| label: fig-1
#### Plot graph on the number of EMS responses in Toronto ####
clean_performance_data |>
  # Define plot 
  ggplot(aes(
    x = year, 
    y = ems_responses
  )) +

  # Define axis labels
  labs(y = "Emergency Medical Responses", x = "Year") +

  # Set the theme
  theme_minimal() +

  # Define styles
  theme(
    axis.title.x = element_text(size = 9, family = "serif", face = "bold"),
    axis.text.x = element_text(size = 8, family = "serif"),
    axis.title.y = element_text(size = 9, family = "serif", face = "bold"),
    axis.text.y = element_text(size = 8, family = "serif")
  ) +

  # Set axis formats
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_x_continuous(breaks = seq(2014, 2021, 1)) +

  # Render bar chart
  geom_bar(stat="identity", fill = "deepskyblue3")
```

\newpage

To further illustrate the responsibility of Toronto's paramedics, @fig-2 outlines the number of incidents, and ambulance dispatches the service fielded in 2021 [@cite_paramedic_incident_data]. The paramedic service responded to over 310,000 calls, ranging from routine medical emergencies to motor vehicle accidents and fires. These calls required a total of 315,788 ambulances, with an average of 865 ambulances dispatched per day. Most calls and incidents need between 1 and three ambulances [@cite_paramedic_incident_data]. However, some severe incidents placed an even greater strain on the service's resources:

-   74 fire emergencies required 5 or more ambulances, with one incident requiring a total of 22 ambulances.
-   46 medical emergencies required 5 or more ambulances, with one incident requiring a total of 11 ambulances.
-   33 motor vehicle accidents required 5 or more ambulances, with one incident requiring a total of 11 ambulances.

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Emergency Medical Incidents and Dispatched Ambulances (in 2021)"
#| label: fig-2
#### Create a table summarizing emergency medical incident data in Toronto ####
summarized_incident_data <- clean_incident_data |>
  # Group data by incident type
  group_by(incident_type) |>

  # Select only incident type and arrived units
  select(incident_type, units_arrived_at_scene) |>

  # Add new columns summarizing data
  summarize(
    # Calculate total number of incident per incident type
    total_incidents = table(incident_type),

    # Calculate total number of ambulances responding to incidents
    units_sum = sum(units_arrived_at_scene, na.rm = TRUE),

    # Calculate the mean of ambulances per incident 
    the_mean = signif(mean(units_arrived_at_scene, na.rm = TRUE), digits = 2),

    # Calculate the mean of ambulances per incident 
    dispatches_day = round(sum(units_arrived_at_scene, na.rm = TRUE)/365, digits = 0), 
  ) |>

  # Rename column headers
  rename("Incident Type" = "incident_type", "Incidents" = "total_incidents", "Dispatches" = "units_sum", "Per Incident" = "the_mean", "Per Day" = "dispatches_day")

# Rename "-" in data to a more meaningful label
summarized_incident_data[1, "Incident Type"] = "Other Incident"

# Order rows alphabetically by incident type
summarized_incident_data <- summarized_incident_data[order(summarized_incident_data$`Incident Type`),]

# Render table
summarized_incident_data |>
  # Table configuration 
  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>

  # Add headers for table data
  add_header_above(
    c(" ", "Year Totals" = 2, "Dispatch Averages" = 2), 
    align = "c",
    bold = TRUE
  ) |>

  # Style the table 
  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>

  # Make the table headers bold
  row_spec(0, bold = TRUE) |>

  # Increase width of first column
  column_spec(1, width = "13em") 
```

\newpage

### Mandated Response Times

In Ontario, the *Ambulance Act* defines the rules governing paramedics' response times. The act defines response times as the difference between the time an emergency service dispatcher receives a call and the on-scene arrival of either an ambulance or a person equipped with a defibrillator [@cite_ambulance_act]. The *Ambulance Act* mandates varying response times based on the severity of a patient's injury. To define these response times and categorize patients based on their injuries, paramedics use the Canadian Triage & Acuity Scale (CTAS). CTAS has 5 levels of severity, with level 1 representing life-threatening injuries and level 5 representing non-urgent injuries [@cite_ctas].

Specific response times are defined under Regulation 257/00, Part VIII of the *Ambulance Act*. This regulation sets the response times for patients suffering from Sudden Cardiac Arrest (SCA) or a CTAS Level 1 injury as 6 minutes and 8 minutes, respectively [@cite_ambulance_act]. CTAS Levels 2 through 5 do not have mandated response times in this regulation, leaving municipalities to define their response times based on the needs of their inhabitants and the abilities of their paramedic service. @fig-3 outlines the City of Toronto's response times for each CTAS level [@cite_ambulance_response_data].

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Mandated Response Times for the Toronto Paramedic Services (as of 2021)"
#| label: fig-3
#### Create a table outlining CTAS levels and response times ####
clean_response_data |>
  # Select type of response and estimated response plan 
  select(patient_category, plan_in_minutes) |>

  # Remove identical rows
  distinct() |>

  # Create new descriptor column 
  cbind(descriptor = c(
    "Sudden Cardiac Arrest", 
    "Resuscitative care", 
    "Emergent care", 
    "Urgent care", 
    "Less urgent care", 
    "Non-urgent care")
  ) |>

  # Rename all columns to human-friendly format
  rename(
    "Response Level" = "patient_category", 
    "Response Time (Mins)" = "plan_in_minutes", 
    "Level Descriptor" = "descriptor"
  ) |>

  # Render table
  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>

  # Style the table 
  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>

  # Make the table headers bold
  row_spec(0, bold = TRUE)

```

It is important to note that not all calls are mandated to receive a response time within these parameters. Individual municipalities are responsible for defining a "performance plan," which determines the percentage of calls that must be covered by their response times [@cite_ambulance_act]. In Toronto, the enacted performance plan requires 75% of all calls to receive a response within the required time frame. However, prior to 2016, only 60% of SCA calls required a response within the mandated time [@cite_ambulance_response_data]. \newpage

### Actual Response Times

The Toronto Paramedic Services has an excellent track record of achieving, and often exceeding, the percentage of calls receiving mandated response times. The graphs below (@fig-4) chart the percentage of incidents served by the mandated response times versus the city's set performance plan [@cite_ambulance_response_data]. To place these graphs in perspective, the City of Toronto's paramedic service averaged a response time of 6 minutes and 58 seconds for "emergency calls" in 2021 [@cite_ambulance_program].

```{r}
#| messages: false
#| echo: false
#| warning: false
#| label: fig-4
#| fig-cap: "Percent of Calls Meeting Mandated Response Times (since 2013)"
#| fig-align: center
#| fig-width: 10
#| fig-height: 5
#### Create graphs plotting mandated and actual response times ####
# Define colors to be used in legend
colors <- c("Actual Response" = "steelblue", "Mandated Response" = "darkred")

clean_response_data |>
  # Plot graphs, define x axis
  ggplot(aes(x = response_time_standard_year)) +

  # Define axis labels
  labs(
    y = "Percent of Calls", 
    x = "Year",
    color = "Legend"
  ) +

  # Set the theme
  theme_minimal() +

  # Define styles for axis, legends, and labels
  theme(
    axis.title.x = element_text(size = 15, family = "serif", face = "bold"),
    axis.text.x = element_text(size = 14, family = "serif", angle = 70, vjust = 0.7),
    axis.title.y = element_text(size = 15, family = "serif", face = "bold"),
    axis.text.y = element_text(size = 14, family = "serif"),
    legend.title = element_text(size = 15, family = "serif", face = "bold"),
    legend.text = element_text(size = 14, family = "serif"),
    strip.text.x = element_text(size = 15, family = "serif", face = "bold"),
    strip.text.y = element_text(size = 14, family = "serif", face = "bold")
  ) +

  # Set axis formats
  scale_x_continuous(breaks=seq(2013, 2021, 2)) +

  # Render lines
  geom_line(aes(y = plan_in_percentage, color = "Mandated Response")) + 
  geom_line(aes(y = performance_in_percentage, color = "Actual Response")) +

  # Set colors for graph; enable legend
  scale_color_manual(values = colors) +

  # Merge graphs into one figure
  facet_wrap(~patient_category) 
```

There was only one year when Toronto's paramedics could not meet all of the requirements set by the city's performance plan. In 2013, only 70% of patients identified as CTAS Level 1 were served by the mandated response time, as opposed to the 75% coverage promised by the performance plan [@cite_ambulance_response_data]. The data provided to the public does not indicate specific response times for each case, so it is impossible to determine how much longer response times were compared to the mandates---in some cases, it may be a difference of a few seconds or more than a few minutes. \newpage

## Discussion

The data used for analysis in this paper was collected and published by the Toronto Paramedic Service. As such, the data is open to institutional bias. The Toronto Paramedic Services determines what data is collected and the methods used for collection. It also has specific goals for data collection: in the case of response times, they must collect the data so they can report their performance to Ontario's Ministry of Health. These goals can impact the data and cause the service to omit significant data points, such as the lack of a measured, case-by-case response time. The data also has other limitations, such as the lack of metrics after 2021.

When analysing a paramedic service's efficacy, an important consideration is whether response times based on the "golden hour" are valuable metrics. In principle, the sooner a paramedic can respond to a call leads to greater efficiency, as it shortens the time paramedics spend on a call and helps patients receive more immediate care. However, healthcare professionals are debating the relevance of the "golden hour" as the foundation of emergency medical services. One study found that "there was no association between EMS intervals and mortality," suggesting that the "golden hour," in the context of EMS, did not impact a patient's chances of survival [@cite_study_1]. A separate study focusing on rural areas found that longer transport times to a trauma center "did not significantly influence survival" [@cite_study_2]. Contrasting these findings is a study from Quebec, which found that spending over 60 minutes before arriving at a trauma center had a "statistically significant" impact on a patient's odds of survival, although this study only surveyed severely injured patients requiring advanced life support (ALS) [@cite_study_3]. These studies reveal that response times based on the "golden hour" principle are of little help in determining a paramedic service's efficacy, except for the most urgent cases that require advanced measures.

With this new limited scope in mind and focusing on only urgent cases, data from the Toronto Paramedic Services reveals that their quick response times effectively improved the chances of survival for critical patients. Patients in Toronto suffering from a stroke face a mortality rate of 11.3%, below the province's average of 12.9% [@cite_stroke_mortality]. Patients in Toronto suffering from a STEMI, or a heart attack caused by a complete blockage of the coronary artery, face a mortality rate of 10.9%, slightly lower than the province's average of 11.2% [@cite_stemi_mortality]. It is important to note that the Toronto Paramedic Service did not include this data in the packages made available on Open Data Toronto.

The Toronto Paramedic Services should include more comprehensive metrics in the data sets they provide to Open Data Toronto. As revealed in this paper, the service's current data sets lack vital information on case-by-case response times and patient outcomes. Where permitted, the service should also include the type of medical emergency addressed by paramedics. The inclusion of these expanded data points would allow for a more meaningful analysis. \newpage

## Inclusive Acknowledgements

**Land Acknowledgement**

We wish to acknowledge the land on which this data was collected and analysed. For thousands of years, it has been the traditional land of the Huron-Wendat, the Seneca, and the Mississaugas of the Credit. Today, this meeting place is still the home to many Indigenous people from across Turtle Island and we are grateful to have the opportunity to work on this land.

This statement was originally prepared in consultation with First Nations House and the Elders Circle for use at the University of Toronto [@cite_land_acknowledgement].\

**Data Acknowledgement**

The data used in this paper was obtained through Open Data Toronto, an initiative to provide openly accessible data about the city [@cite_about_opendatatoronto]. This paper uses three data sets from Open Data Toronto, each provided by the Toronto Paramedic Services:

-   Paramedic Services Incident Data[^1]
-   Land Ambulance Response Time Standard[^2]
-   Pre-Hospital Emergency Care Performance Metrics[^3]

[^1]: Paramedic Services Incident Data is available at: https://open.toronto.ca/dataset/paramedic-services-incident-data

[^2]: Land Ambulance Response Time Standard is available at: https://open.toronto.ca/dataset/land-ambulance-response-time-standard

[^3]: Pre-Hospital Emergency Care Performance Metrics is available at: https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics

This project contains information licensed under the Open Government Licence -- Toronto[^4].\

[^4]: Open Government License -- Toronto is available at: https://open.toronto.ca/open-data-license/

**Resources Acknowledgment**

The primary tool used to analyse data in this paper is R, an open-source statistical programming language [@cite_tool_r]. The paper also uses a number of R packages, including: dplyr [@cite_tool_dplyr], ggplot2 [@cite_tool_ggplot2], here [@cite_tool_here], janitor [@cite_tool_janitor], kableExtra [@cite_tool_kableextra], knitr [@cite_tool_knitr], lubridate [@cite_tool_lubridate], opendatatoronto [@cite_tool_opendatatoronto], readr [@cite_tool_readr], RColorBrewer [@cite_tool_rcolorbrewer], scales [@cite_tool_scales], and tidyverse [@cite_tool_tidyverse]. \newpage

## References

<!-- Comment required to show references as a numbered section -->

### 2023-03-11 scripts/00-install_dependencies.qmd
Init

---
title: "Install the dependencies required for this project"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Preamble ####
# Purpose: Install the dependencies required for this project
# Prerequisites: You must also have the R language installed
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses
```

```{r}
# Install required packages (you only need to do this once)
install.packages("lubridate")
install.packages("tidyverse")
install.packages("dplyr")
install.packages("knitr")
install.packages("janitor")
install.packages("scales")
install.packages("RColorBrewer")
install.packages("ggplot2")
install.packages("kableExtra")
install.packages("here")
install.packages("readr")
```

### 2023-03-11 scripts/01-data_import.qmd
Init

---
title: "Import data from the Toronto Paramedic Service via Open Data Toronto"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Preamble ####
# Purpose: Import data from the Toronto Paramedic Service via Open Data Toronto
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses

#### Workspace set-up ####
library(opendatatoronto)
library(tidyverse)
library(dplyr)
```

```{r}
#| warning: false
#### Fetch "Land Ambulance Response Time Standard" from Open Data Toronto
# Link: https://open.toronto.ca/dataset/land-ambulance-response-time-standard 
raw_response_data <-
  list_package_resources("8b068db1-95b8-4ea9-b942-8a7ce3a5fbd6") |>
  filter(name == "Response Time (2013 - 2021)") |>
  get_resource()

# Write data to a .csv file
write_csv(
  x = raw_response_data$`Land Ambulance Services`,
  file = "../inputs/data/raw_response_data.csv"
)
```

```{r}
#| warning: false
#### Fetch "Pre-Hospital Emergency Care Performance Metrics" from Open Data Toronto ####
# Link: https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
raw_performance_data <-
  list_package_resources("174ace54-bcc7-4885-85c1-76094ab4a0ca") |>
  filter(name == "Pre-hospital emergency care performance metrics 2014-2021") |>
  get_resource() |>
  filter(`Pre-Hospital Emergency Care Type` == "Emergency Request for Service")

# Write data to a .csv file
write_csv(
  x = raw_performance_data,
  file = "../inputs/data/raw_performance_data.csv"
)
```

```{r}
#| warning: false
#### Fetch "Paramedic Services Incident Data" from Open Data Toronto ####
# Link: https://open.toronto.ca/dataset/paramedic-services-incident-data
raw_incident_data <-
  list_package_resources("c21f3bd1-e016-4469-abf5-c58bb8e8b5ce") |>
  filter(name == "paramedic-services-incident-data-2017-2021") |>
  get_resource()

# Write data to a .csv file
write_csv(
  x = raw_incident_data$`2021`,
  file = "../inputs/data/raw_incident_data.csv",
)

# Note: creating this .csv may take some time as it's a large data set
```

### 2023-03-11 scripts/02-data_cleaning.qmd
Init

---
title: "Import raw csv data, clean it, and generate new csv files"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Preamble ####
# Purpose: Import raw csv data, clean it, and generate new csv files
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses

#### Workspace set-up ####
library(tidyverse)
library(dplyr)
library(here)
library(readr)
library(janitor)
```

```{r}
#### Fetch and clean the raw response data ####
raw_happy_data <- readr::read_csv(here::here("inputs/data/raw_data.csv"), show_col_types = FALSE) |>
  # Change column naming format
  clean_names() |> 
  select(year, happy) |>
  filter(happy >= 2) |> 
  count(year, happy) |>
  group_by(year) |>
  summarise(people = sum(n))

raw_directions_data <- readr::read_csv(here::here("inputs/data/raw_data.csv"), show_col_types = FALSE) |>
  # Change column naming format
  clean_names() |> 
  select(year, directns) |>
  filter(directns >= 1, directns <= 4) |> 
  count(year, directns) |>
  group_by(year) |>
  summarise(people = sum(n))

raw_give_homeless_data <- readr::read_csv(here::here("inputs/data/raw_data.csv"), show_col_types = FALSE) |>
  # Change column naming format
  clean_names() |> 
  select(year, givhmlss) |>
  filter(givhmlss >= 1, givhmlss <= 4) |>
  count(year, givhmlss) |>
  group_by(year) |>
  summarise(people = sum(n))

colors <- c(
  "Happiness" = "steelblue", 
  "Give Directions to Stranger" = "darkred",
  "Give Money to Homeless" = "darkgreen")

ggplot() +
  # Define axis labels
  labs(
    y = "People", 
    x = "Year",
    color = "Legend"
  ) +

  # Set the theme
  theme_minimal() +

  geom_line(data = raw_happy_data, aes(x = year, y = people, color = "Happiness")) +
  geom_line(data = raw_directions_data, aes(x = year, y = people, color = "Give Directions to Stranger")) +
  geom_line(data = raw_give_homeless_data, aes(x = year, y = people, color = "Give Money to Homeless")) +
  scale_x_continuous(breaks=c(2002,2004,2012,2014)) +
  scale_color_manual(values = colors)

```

```{r}
#### Fetch and clean the raw response data ####
raw_happy_data <- readr::read_csv(here::here("inputs/data/raw_data.csv"), show_col_types = FALSE) |>
  # Change column naming format
  clean_names() |> 
  select(year, happy) |>
  filter(happy == 3) |> 
  count(year, happy) |>
  group_by(year) |>
  summarise(people = sum(n))

raw_directions_data <- readr::read_csv(here::here("inputs/data/raw_data.csv"), show_col_types = FALSE) |>
  # Change column naming format
  clean_names() |> 
  select(year, directns) |>
  filter(directns >= 5) |> 
  count(year, directns) |>
  group_by(year) |>
  summarise(people = sum(n))

raw_give_homeless_data <- readr::read_csv(here::here("inputs/data/raw_data.csv"), show_col_types = FALSE) |>
  # Change column naming format
  clean_names() |> 
  select(year, givhmlss) |>
  filter(givhmlss >= 5) |>
  count(year, givhmlss) |>
  group_by(year) |>
  summarise(people = sum(n))

colors <- c(
  "Happiness" = "steelblue", 
  "Give Directions to Stranger" = "darkred",
  "Give Money to Homeless" = "darkgreen")

ggplot() +
  # Define axis labels
  labs(
    y = "People", 
    x = "Year",
    color = "Legend"
  ) +

  # Set the theme
  theme_minimal() +

  geom_line(data = raw_happy_data, aes(x = year, y = people, color = "Happiness")) +
  geom_line(data = raw_directions_data, aes(x = year, y = people, color = "Give Directions to Stranger")) +
  geom_line(data = raw_give_homeless_data, aes(x = year, y = people, color = "Give Money to Homeless")) +
  scale_x_continuous(breaks=c(2002,2004,2012,2014)) +
  scale_color_manual(values = colors)

```

### 2023-03-11 scripts/03-data_testing.qmd
Init

---
title: "Testing Data from the Toronto Paramedic Service"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Preamble ####
# Purpose: Testing Data from the Toronto Paramedic Service
# Prerequisites: Need to have downloaded the data using [inputs/data/data_import.qmd]
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses

#### Workspace set-up ####
library(tidyverse)
library(janitor)
library(here)
library(readr)
```

```{r}
#### Test the clean response data #### 
# Fetch the clean response data 
clean_response_data <- readr::read_csv(here::here("inputs/data/clean_response_data.csv"), show_col_types = FALSE)

# Test that the target performance is less than or equal to 75%
clean_response_data$plan_in_percentage |> max() <= 0.75

# Test that the reported year is greater than or equal to 2013
clean_response_data$response_time_standard_year |> min() >= 2013

# Test that the actual performance data is less than or equal to 100%
clean_response_data$performance_in_percentage |> max() <= 1.00

# Test that the actual performance data is greater than or equal to 0%
clean_response_data$performance_in_percentage |> min() >= 0.00
```

```{r}
#### Test the clean performance data #### 
# Fetch the clean performance data
clean_performance_data <- readr::read_csv(here::here("inputs/data/clean_performance_data.csv"), show_col_types = FALSE)

# Test that the reported year is greater than or equal to 2014
clean_performance_data$year |> min() >= 2014

# Test that EMS responses is a number
clean_performance_data$ems_responses |> class() == "numeric"

# Test that EMS responses is not greater than Toronto's population
clean_performance_data$ems_responses |> max() <= 3000000

# Test that EMS responses is not 0 
clean_performance_data$ems_responses |> min() >= 0
```

```{r}
#### Test the clean incident data #### 
# Fetch the clean incident data
clean_incident_data <- readr::read_csv(here::here("inputs/data/clean_incident_data.csv"), show_col_types = FALSE)

# Test that the units arrived on scene is not greater than the city's number of paramedics
clean_incident_data$units_arrived_at_scene |> max() <= 1186

# Test that the units arrive on scene is greater than zero
clean_incident_data$units_arrived_at_scene |> min() >= 0

# Test that the dispatch time is a date 
clean_incident_data$dispatch_time |> class() == "POSIXct"

# Test that the incident type contains is comprised of characters
clean_incident_data$incident_type |> class() == "character"

# Note: due to an issue with the dispatch_time's column formatting as a date,
# the second returned Boolean may be false. As long as the first value returns 
# true, the data passes the test.
```

### 2023-03-11 scripts/04-data_simulation.qmd
Init

---
title: "Simulating Data from the Toronto Paramedic Service"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Preamble ####
# Purpose: Read in data from the Toronto Paramedic Service
# Prerequisites: Need to know where to get Toronto Paramedic Services data
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses

#### Workspace set-up ####
library(tidyverse) # A collection of data-related packages
library(janitor) # Helps clean datasets
```

```{r}
#### Simulate data based on Land Ambulance Response Time Standard ####
# Link: https://open.toronto.ca/dataset/land-ambulance-response-time-standard 
simulated_response_data <-
  tibble(
    # Patient categories, predefined and set terms
    "patient_category" = c(
      "SCA", 
      "CTAS 1",
      "CTAS 2",
      "CTAS 3",
      "CTAS 4",
      "CTAS 5"
    ),
    # Response time mandates, predefined and set numbers
    "plan_in_minutes" = c(
      6, 
      8,
      10,
      15,
      20,
      25
    ),
    # Performance mandates, predefined and set percentage
    "plan_in_percentage" = sample(
      x = 0.75,
      size = 6,
      replace = TRUE
    ),
    # Actual percentage of response times meeting performance plan, varying between 60% and 99%
    "performance_in_percentage" = sample(
      seq(0.6, 0.99, by = 0.001),
      size = 6,
      replace = TRUE
    ),
    # Sample year associated with data
    "response_time_standard_tear" = c(
      2021, 
      2021,
      2021,
      2021,
      2021,
      2021
    ),
)

simulated_response_data
```

```{r}
#### Simulate data based on Pre-Hospital Emergency Care Performance Metrics ####
# Link: https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
simulated_performance_data <-
  tibble(
    # List all years between 2014 and 2021, inclusive
    "year" = c(2014:2021),

    # Randomly select numbers between 200,000 and 300,000
    "ems_responses" = sample(
      x = c(200000:350000),
      size = 8,
      replace = TRUE
    )
)

simulated_performance_data
```

```{r}
#### Simulate data based on Paramedic Services Incident Data ####
# Link: https://open.toronto.ca/dataset/paramedic-services-incident-data
simulated_incident_data <-
  tibble(
    # Randomly pick dates between January 2, 2021 and December 31, 2021
    "dispatch_time" = sample(
      seq(as.Date('2021/01/01'), as.Date('2021/12/31'), by="day"), 
      size = 1000,
      replace = TRUE
    ),

    # Randomly pick an incident type 1000 times
    "incident_type" = sample(
      x = c(
        "Medical",
        "Motor Vehicle Accident",
        "Emergency Transfer",
        "Fire",
        "Airport Standby",
        "Other"
      ),
      size = 1000,
      replace = TRUE
    ),

    # Randomly select the number of units arriving on scene, between 1 and 22
    "units_arrived_at_scene" = sample(
      x = c(1:22),
      size = 1000,
      replace = TRUE
    )
  )

simulated_incident_data
```

### 2023-03-11 inputs/data/gss-norc/GSS.r

@@ -1,0 +1,45 @@
+library(foreign)
+library(here)
+library(readr)
+library(dplyr)
+
+  read.dct <- function(dct, labels.included = "yes") {
+      temp <- readLines(dct)
+      temp <- temp[grepl("_column", temp)]
+      switch(labels.included,
+             yes = {
+                 pattern <- "_column\\(([0-9]+)\\)\\s+([a-z0-9]+)\\s+(.*)\\s+%([0-9]+)[a-z]\\s+(.*)"
+                 classes <- c("numeric", "character", "character", "numeric", "character")
+                 N <- 5
+                 NAMES <- c("StartPos", "Str", "ColName", "ColWidth", "ColLabel")
+             },
+             no = {
+                 pattern <- "_column\\(([0-9]+)\\)\\s+([a-z0-9]+)\\s+(.*)\\s+%([0-9]+).*"
+                 classes <- c("numeric", "character", "character", "numeric")
+                 N <- 4
+                 NAMES <- c("StartPos", "Str", "ColName", "ColWidth")
+             })
+      temp_metadata <- setNames(lapply(1:N, function(x) {
+          out <- gsub(pattern, paste("\\", x, sep = ""), temp)
+          out <- gsub("^\\s+|\\s+$", "", out)
+          out <- gsub('\"', "", out, fixed = TRUE)
+          class(out) <- classes[x] ; out }), NAMES)
+      temp_metadata[["ColName"]] <- make.names(gsub("\\s", "", temp_metadata[["ColName"]]))
+      temp_metadata
+  }
+
+  read.dat <- function(dat, metadata_var, labels.included = "yes") {
+      read.fwf(dat, widths = metadata_var[["ColWidth"]], sep = "", col.names = metadata_var[["ColName"]])
+  }
+
+
+GSS_metadata <- read.dct(here::here("inputs/data/gss-norc/GSS.dct"))
+GSS_ascii <- read.dat(here::here("inputs/data/gss-norc/GSS.dat"), GSS_metadata)
+attr(GSS_ascii, "col.label") <- GSS_metadata[["ColLabel"]]
+GSS <- GSS_ascii |>
+  select(BALLOT, HAPPY, GIVHMLSS, DIRECTNS, YEAR)
+
+write_csv(
+  x = GSS,
+  file = here::here("inputs/data/raw_data.csv"),
+)

### 2023-03-12 outputs/paper/paper.qmd
Updates to file structure and content

@@
 ---
-title: "The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?"
+title: "Does Happiness Lead to Altruism: A review of US Generial Social Survey data"
 format: pdf
-author: "Sebastian Rodriguez^[University of Toronto, Faculty of Information]"
-abstract: "The Toronto Paramedic Services provides all pre-hospital care to the city’s inhabitants.
-Recent criticism of the service called into question its performance, highlighting staff shortages and institutional challenges that the city's paramedics face.
-This paper investigates the Toronto Paramedic Services’ efficacy by comparing its emergency response times to the standards set by provincial and municipal governments.
-It reveals that Toronto's paramedics meet and often exceed response time standards, and provide stroke and STEMI heart attack patients with a chance of survival greater than the provincial average."
+author: "Sebastian Rodriguez, Iz Leitch, Randall Ni"
+abstract: "Something about happiness and altruism" 
 date: today
@@
 bibliography: references.bib
-thanks: "Code and data supporting this analysis is available at: https://github.com/seb646/toronto-paramedic-responses"
+thanks: "Code and data supporting this analysis is available at: https://github.com/seb646/happiness-and-altruism"
 ---
@@
 # Fetch the the clean response data
-clean_response_data <- readr::read_csv(here::here("inputs/data/clean_response_data.csv"), show_col_types = FALSE)
+happy_positive_data <- readr::read_csv(here::here("inputs/data/clean/happy_positive_data.csv"), show_col_types = FALSE)
 
 # Fetch clean performance data
-clean_performance_data <- readr::read_csv(here::here("inputs/data/clean_performance_data.csv"), show_col_types = FALSE)
+directions_positive_data <- readr::read_csv(here::here("inputs/data/clean/directions_positive_data.csv"), show_col_types = FALSE)
 
 # Fetch the clean incident response
-clean_incident_data <- readr::read_csv(here::here("inputs/data/clean_incident_data.csv"), show_col_types = FALSE)
+homeless_positive_data <- readr::read_csv(here::here("inputs/data/clean/homeless_positive_data.csv"), show_col_types = FALSE)
+
+# Fetch the the clean response data
+happy_negative_data <- readr::read_csv(here::here("inputs/data/clean/happy_negative_data.csv"), show_col_types = FALSE)
+
+# Fetch clean performance data
+directions_negative_data <- readr::read_csv(here::here("inputs/data/clean/directions_negative_data.csv"), show_col_types = FALSE)
+
+# Fetch the clean incident response
+homeless_negative_data <- readr::read_csv(here::here("inputs/data/clean/homeless_negative_data.csv"), show_col_types = FALSE)
 ```
@@
 
-A city's emergency medical services (EMS) are integral to its health care system.
-The paramedics and emergency medical technicians working for the service provide pre-hospital assessment, stabilisation, and care to people needing medical assistance.
-These professionals respond to every call from the public for aid, from broken bones and heart attacks to car accidents and house fires.
-Current emergency medical services operate on the principle of the "golden hour," which is the 60 minutes between an initial injury and arrival at a trauma center when a patient has the highest statistical chance of survival [@cite_golden_hour].
-This "golden hour" is the foundation of all emergency medical responses and requires medical personnel to arrive on scene in under 10 minutes [@cite_golden_hour_10_mins].
-
-In Toronto, inhabitants receive emergency medical assistance through the Toronto Paramedic Services.
-As of 2020, the Toronto Paramedic Services employed 1,186 paramedics operating 234 ambulances out of 45 stations across the city [@cite_toronto_paramedic_2020_report].
-The service provides all land ambulatory responses within the city, including on-scene treatment and stabilisation of patients, transport to a hospital, and transport between hospitals.
-The Province of Ontario and the City of Toronto set specific requirements for the length of time an ambulance should take when responding to an emergency call, which can be anywhere between 6 to 25 minutes.
-While the service strives to achieve these mandated goals, they face staffing, budget, and other institutional challenges that impact its ability to provide care.
-Recently, the service gained attention for slower response times because of staffing shortages in hospitals and in the service itself [@cite_paramedic_shortages].
-These staffing shortages decrease the number of available paramedics while simultaneously increasing the time it takes for a patient to enter a hospital, tying up paramedics at hospitals.
-The recent COVID-19 pandemic also impacted the service, with most paramedics stating that morale is "either worse or as bad as it has ever been" [@cite_paramedic_morale].
-
-Despite the challenges that Toronto's paramedics face, they must respond to emergencies quickly to provide patients with the best chances of survival.
-This paper investigates the response times of the Toronto Paramedic Service compared to the mandated requirements set by the local and provincial governments.
-Specifically, the paper aims to place response times within a larger context of the service's responsibility and interrogate the relevance of the "golden hour" principle to determine the efficacy of the service.
-
-The findings indicated a growing trend in the number of emergency medical calls, an increase of approximately 10,000 per year since 2014 [@cite_emergency_care_data].
-Despite this increase, the Toronto Paramedic Services met and often exceeded the response times set by both the Province of Ontario and the City of Toronto.
-The paper also revealed a significant lack of paramedic metrics from the data sets available on Open Data Toronto.
-Without metrics on case-by-case response times and patient outcomes, other sources were required to inform data analysis.
-Overall, the response time of Toronto's paramedics contributed to stroke and mortality rates lower than Ontario's average, giving the city's inhabitants greater chances of survival when suffering from life-threatening injuries.
 \newpage
 
-## Paramedic Response Data
-
-### Scope of Responsibility
-
-Before investigating the response times and efficacy of the Toronto Paramedic Services, it is important to understand the service's breadth of responsibility.
-With a population of nearly 3 million people, Toronto requires a medical response to roughly 300,000 emergency calls every year.
-As the city's population grows, so too does the need for medical assistance.
-There was a steady increase of approximately 10,000 emergency medical requests per year from 2014 to 2019, as outlined by [@fig-1; @cite_emergency_care_data].
-This trend ended briefly in 2020 when the COVID-19 pandemic hit Toronto, placing an incredible strain on the city's healthcare system.
-Due to a concern about COVID-19 infection, many people delayed or avoided medical care throughout the pandemic [@cite_avoiding_hospitals].
+## General Social Survey Data
 
@@
 #| warning: false
-#| fig-cap: "Number of Emergency Medical Responses in Toronto (by year since 2014)"
+#| fig-cap: "Happiness vs Altruism"
 #| label: fig-1
-#### Plot graph on the number of EMS responses in Toronto ####
-clean_performance_data |>
-  # Define plot 
-  ggplot(aes(
-    x = year, 
-    y = ems_responses
-  )) +
 
-  # Define axis labels
-  labs(y = "Emergency Medical Responses", x = "Year") +
+colors <- c(
+  "Happiness" = "steelblue",
+  "Give Directions to Stranger" = "darkred",
+  "Give Money to Homeless" = "darkgreen"
+)
 
-  # Set the theme
-  theme_minimal() +
-
-  # Define styles
-  theme(
-    axis.title.x = element_text(size = 9, family = "serif", face = "bold"),
-    axis.text.x = element_text(size = 8, family = "serif"),
-    axis.title.y = element_text(size = 9, family = "serif", face = "bold"),
-    axis.text.y = element_text(size = 8, family = "serif")
-  ) +
-
-  # Set axis formats
-  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
-  scale_x_continuous(breaks = seq(2014, 2021, 1)) +
-
-  # Render bar chart
-  geom_bar(stat="identity", fill = "deepskyblue3")
-```
-
-\newpage
-
-To further illustrate the responsibility of Toronto's paramedics, @fig-2 outlines the number of incidents, and ambulance dispatches the service fielded in 2021 [@cite_paramedic_incident_data].
-The paramedic service responded to over 310,000 calls, ranging from routine medical emergencies to motor vehicle accidents and fires.
-These calls required a total of 315,788 ambulances, with an average of 865 ambulances dispatched per day.
-Most calls and incidents need between 1 and three ambulances [@cite_paramedic_incident_data].
-However, some severe incidents placed an even greater strain on the service's resources:
-
--   74 fire emergencies required 5 or more ambulances, with one incident requiring a total of 22 ambulances.
--   46 medical emergencies required 5 or more ambulances, with one incident requiring a total of 11 ambulances.
--   33 motor vehicle accidents required 5 or more ambulances, with one incident requiring a total of 11 ambulances.
-
-```{r}
-#| messages: false
-#| echo: false
-#| warning: false
-#| fig-cap: "Emergency Medical Incidents and Dispatched Ambulances (in 2021)"
-#| label: fig-2
-#### Create a table summarizing emergency medical incident data in Toronto ####
-summarized_incident_data <- clean_incident_data |>
-  # Group data by incident type
-  group_by(incident_type) |>
-
-  # Select only incident type and arrived units
-  select(incident_type, units_arrived_at_scene) |>
-
-  # Add new columns summarizing data
-  summarize(
-    # Calculate total number of incident per incident type
-    total_incidents = table(incident_type),
-
-    # Calculate total number of ambulances responding to incidents
-    units_sum = sum(units_arrived_at_scene, na.rm = TRUE),
-
-    # Calculate the mean of ambulances per incident 
-    the_mean = signif(mean(units_arrived_at_scene, na.rm = TRUE), digits = 2),
-
-    # Calculate the mean of ambulances per incident 
-    dispatches_day = round(sum(units_arrived_at_scene, na.rm = TRUE)/365, digits = 0), 
-  ) |>
-
-  # Rename column headers
-  rename("Incident Type" = "incident_type", "Incidents" = "total_incidents", "Dispatches" = "units_sum", "Per Incident" = "the_mean", "Per Day" = "dispatches_day")
-
-# Rename "-" in data to a more meaningful label
-summarized_incident_data[1, "Incident Type"] = "Other Incident"
-
-# Order rows alphabetically by incident type
-summarized_incident_data <- summarized_incident_data[order(summarized_incident_data$`Incident Type`),]
-
-# Render table
-summarized_incident_data |>
-  # Table configuration 
-  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
-
-  # Add headers for table data
-  add_header_above(
-    c(" ", "Year Totals" = 2, "Dispatch Averages" = 2), 
-    align = "c",
-    bold = TRUE
-  ) |>
-
-  # Style the table 
-  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
-
-  # Make the table headers bold
-  row_spec(0, bold = TRUE) |>
-
-  # Increase width of first column
-  column_spec(1, width = "13em") 
-```
-
-\newpage
-
-### Mandated Response Times
-
-In Ontario, the *Ambulance Act* defines the rules governing paramedics' response times.
-The act defines response times as the difference between the time an emergency service dispatcher receives a call and the on-scene arrival of either an ambulance or a person equipped with a defibrillator [@cite_ambulance_act].
-The *Ambulance Act* mandates varying response times based on the severity of a patient's injury.
-To define these response times and categorize patients based on their injuries, paramedics use the Canadian Triage & Acuity Scale (CTAS).
-CTAS has 5 levels of severity, with level 1 representing life-threatening injuries and level 5 representing non-urgent injuries [@cite_ctas].
-
-Specific response times are defined under Regulation 257/00, Part VIII of the *Ambulance Act*.
-This regulation sets the response times for patients suffering from Sudden Cardiac Arrest (SCA) or a CTAS Level 1 injury as 6 minutes and 8 minutes, respectively [@cite_ambulance_act].
-CTAS Levels 2 through 5 do not have mandated response times in this regulation, leaving municipalities to define their response times based on the needs of their inhabitants and the abilities of their paramedic service.
-@fig-3 outlines the City of Toronto's response times for each CTAS level [@cite_ambulance_response_data].
-
-```{r}
-#| messages: false
-#| echo: false
-#| warning: false
-#| fig-cap: "Mandated Response Times for the Toronto Paramedic Services (as of 2021)"
-#| label: fig-3
-#### Create a table outlining CTAS levels and response times ####
-clean_response_data |>
-  # Select type of response and estimated response plan 
-  select(patient_category, plan_in_minutes) |>
-
-  # Remove identical rows
-  distinct() |>
-
-  # Create new descriptor column 
-  cbind(descriptor = c(
-    "Sudden Cardiac Arrest", 
-    "Resuscitative care", 
-    "Emergent care", 
-    "Urgent care", 
-    "Less urgent care", 
-    "Non-urgent care")
-  ) |>
-
-  # Rename all columns to human-friendly format
-  rename(
-    "Response Level" = "patient_category", 
-    "Response Time (Mins)" = "plan_in_minutes", 
-    "Level Descriptor" = "descriptor"
-  ) |>
-
-  # Render table
-  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
-
-  # Style the table 
-  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
-
-  # Make the table headers bold
-  row_spec(0, bold = TRUE)
-
-```
-
-It is important to note that not all calls are mandated to receive a response time within these parameters.
-Individual municipalities are responsible for defining a "performance plan," which determines the percentage of calls that must be covered by their response times [@cite_ambulance_act].
-In Toronto, the enacted performance plan requires 75% of all calls to receive a response within the required time frame.
-However, prior to 2016, only 60% of SCA calls required a response within the mandated time [@cite_ambulance_response_data].
-\newpage
-
-### Actual Response Times
-
-The Toronto Paramedic Services has an excellent track record of achieving, and often exceeding, the percentage of calls receiving mandated response times.
-The graphs below (@fig-4) chart the percentage of incidents served by the mandated response times versus the city's set performance plan [@cite_ambulance_response_data].
-To place these graphs in perspective, the City of Toronto's paramedic service averaged a response time of 6 minutes and 58 seconds for "emergency calls" in 2021 [@cite_ambulance_program].
-
-```{r}
-#| messages: false
-#| echo: false
-#| warning: false
-#| label: fig-4
-#| fig-cap: "Percent of Calls Meeting Mandated Response Times (since 2013)"
-#| fig-align: center
-#| fig-width: 10
-#| fig-height: 5
-#### Create graphs plotting mandated and actual response times ####
-# Define colors to be used in legend
-colors <- c("Actual Response" = "steelblue", "Mandated Response" = "darkred")
-
-clean_response_data |>
-  # Plot graphs, define x axis
-  ggplot(aes(x = response_time_standard_year)) +
-
+ggplot() +
   # Define axis labels
   labs(
-    y = "Percent of Calls", 
+    y = "People",
     x = "Year",
@@
 
-  # Define styles for axis, legends, and labels
-  theme(
-    axis.title.x = element_text(size = 15, family = "serif", face = "bold"),
-    axis.text.x = element_text(size = 14, family = "serif", angle = 70, vjust = 0.7),
-    axis.title.y = element_text(size = 15, family = "serif", face = "bold"),
-    axis.text.y = element_text(size = 14, family = "serif"),
-    legend.title = element_text(size = 15, family = "serif", face = "bold"),
-    legend.text = element_text(size = 14, family = "serif"),
-    strip.text.x = element_text(size = 15, family = "serif", face = "bold"),
-    strip.text.y = element_text(size = 14, family = "serif", face = "bold")
+  geom_line(data = happy_positive_data, aes(x = year, y = people, color = "Happiness")) +
+  geom_line(data = directions_positive_data, aes(x = year, y = people, color = "Give Directions to Stranger")) +
+  geom_line(data = homeless_positive_data, aes(x = year, y = people, color = "Give Money to Homeless")) +
+  scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
+  scale_color_manual(values = colors)
+```
+
+```{r}
+#| messages: false
+#| echo: false
+#| warning: false
+#| fig-cap: "Unhappiness vs. No Altruism"
+#| label: fig-2
+
+colors <- c(
+  "Happiness" = "steelblue",
+  "Give Directions to Stranger" = "darkred",
+  "Give Money to Homeless" = "darkgreen"
+)
+
+ggplot() +
+  # Define axis labels
+  labs(
+    y = "People",
+    x = "Year",
+    color = "Legend"
   ) +
 
-  # Set axis formats
-  scale_x_continuous(breaks=seq(2013, 2021, 2)) +
+  # Set the theme
+  theme_minimal() +
 
-  # Render lines
-  geom_line(aes(y = plan_in_percentage, color = "Mandated Response")) + 
-  geom_line(aes(y = performance_in_percentage, color = "Actual Response")) +
-
-  # Set colors for graph; enable legend
-  scale_color_manual(values = colors) +
-
-  # Merge graphs into one figure
-  facet_wrap(~patient_category) 
+  geom_line(data = happy_negative_data, aes(x = year, y = people, color = "Happiness")) +
+  geom_line(data = directions_negative_data, aes(x = year, y = people, color = "Give Directions to Stranger")) +
+  geom_line(data = homeless_negative_data, aes(x = year, y = people, color = "Give Money to Homeless")) +
+  scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
+  scale_color_manual(values = colors)
 ```
-
-There was only one year when Toronto's paramedics could not meet all of the requirements set by the city's performance plan.
-In 2013, only 70% of patients identified as CTAS Level 1 were served by the mandated response time, as opposed to the 75% coverage promised by the performance plan [@cite_ambulance_response_data].
-The data provided to the public does not indicate specific response times for each case, so it is impossible to determine how much longer response times were compared to the mandates---in some cases, it may be a difference of a few seconds or more than a few minutes.
-\newpage
-
 ## Discussion
 
-The data used for analysis in this paper was collected and published by the Toronto Paramedic Service.
-As such, the data is open to institutional bias.
-The Toronto Paramedic Services determines what data is collected and the methods used for collection.
-It also has specific goals for data collection: in the case of response times, they must collect the data so they can report their performance to Ontario's Ministry of Health.
-These goals can impact the data and cause the service to omit significant data points, such as the lack of a measured, case-by-case response time.
-The data also has other limitations, such as the lack of metrics after 2021.
-
-When analysing a paramedic service's efficacy, an important consideration is whether response times based on the "golden hour" are valuable metrics.
-In principle, the sooner a paramedic can respond to a call leads to greater efficiency, as it shortens the time paramedics spend on a call and helps patients receive more immediate care.
-However, healthcare professionals are debating the relevance of the "golden hour" as the foundation of emergency medical services.
-One study found that "there was no association between EMS intervals and mortality," suggesting that the "golden hour," in the context of EMS, did not impact a patient's chances of survival [@cite_study_1].
-A separate study focusing on rural areas found that longer transport times to a trauma center "did not significantly influence survival" [@cite_study_2].
-Contrasting these findings is a study from Quebec, which found that spending over 60 minutes before arriving at a trauma center had a "statistically significant" impact on a patient's odds of survival, although this study only surveyed severely injured patients requiring advanced life support (ALS) [@cite_study_3].
-These studies reveal that response times based on the "golden hour" principle are of little help in determining a paramedic service's efficacy, except for the most urgent cases that require advanced measures.
-
-With this new limited scope in mind and focusing on only urgent cases, data from the Toronto Paramedic Services reveals that their quick response times effectively improved the chances of survival for critical patients.
-Patients in Toronto suffering from a stroke face a mortality rate of 11.3%, below the province's average of 12.9% [@cite_stroke_mortality].
-Patients in Toronto suffering from a STEMI, or a heart attack caused by a complete blockage of the coronary artery, face a mortality rate of 10.9%, slightly lower than the province's average of 11.2% [@cite_stemi_mortality].
-It is important to note that the Toronto Paramedic Service did not include this data in the packages made available on Open Data Toronto.
-
-The Toronto Paramedic Services should include more comprehensive metrics in the data sets they provide to Open Data Toronto.
-As revealed in this paper, the service's current data sets lack vital information on case-by-case response times and patient outcomes.
-Where permitted, the service should also include the type of medical emergency addressed by paramedics.
-The inclusion of these expanded data points would allow for a more meaningful analysis.
 \newpage
@@
 
-The data used in this paper was obtained through Open Data Toronto, an initiative to provide openly accessible data about the city [@cite_about_opendatatoronto].
-This paper uses three data sets from Open Data Toronto, each provided by the Toronto Paramedic Services:
+The data used in this paper was obtained through the General Social Survey (GSS), a project of the independent research organization NORC at the University of Chicago, with principal funding from the National Science Foundation.
+This paper uses three data sets from the NORC's GSS:
 
--   Paramedic Services Incident Data[^1]
--   Land Ambulance Response Time Standard[^2]
--   Pre-Hospital Emergency Care Performance Metrics[^3]
+-   General Happiness[^1]
+-   Has Given Directions to a Stranger[^2]
+-   Has Given Money or Food to a Homeless Person[^3]
 
-[^1]: Paramedic Services Incident Data is available at: https://open.toronto.ca/dataset/paramedic-services-incident-data
+[^1]: General Happiness data is available at: https://gssdataexplorer.norc.org/variables/434/vshow
 
-[^2]: Land Ambulance Response Time Standard is available at: https://open.toronto.ca/dataset/land-ambulance-response-time-standard
+[^2]: Has Given Directions to a Stranger data is available at: https://gssdataexplorer.norc.org/variables/2886/vshow
 
-[^3]: Pre-Hospital Emergency Care Performance Metrics is available at: https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
+[^3]: Has Given Money or Food to a Homeless Person data is available at: https://gssdataexplorer.norc.org/variables/2878/vshow
 
-This project contains information licensed under the Open Government Licence -- Toronto[^4].\
-
-[^4]: Open Government License -- Toronto is available at: https://open.toronto.ca/open-data-license/
+The General Social Survey (GSS) is a project of the independent research organization NORC at the University of Chicago, with principal funding from the National Science Foundation.
 

### 2023-03-12 scripts/00-install_dependencies.qmd
Updates to file structure and content

@@
 install.packages("readr")
+install.packages("foreign")
 ```

### 2023-03-12 scripts/01-data_convert.r

@@ -1,9 +1,20 @@
+#### Preamble ####
+# Purpose: Import data from the Toronto Paramedic Service via Open Data Toronto
+# Author: Sebastian Rodriguez
+# Email: me@srod.ca
+# Date: 3 February 2023
+# GitHub: https://github.com/seb646/toronto-paramedic-responses
+
+#### Workspace set-up ####
 library(foreign)
+library(tidyverse)
+library(dplyr)
 library(here)
 library(readr)
-library(dplyr)
 
-  read.dct <- function(dct, labels.included = "yes") {
+#### Convert raw data from .dat/.dct to .csv ####
+#### Code originally from NORC GSS, modified 
+read.dct <- function(dct, labels.included = "yes") {
       temp <- readLines(dct)
       temp <- temp[grepl("_column", temp)]
       switch(labels.included,
@@ -26,20 +37,20 @@ library(dplyr)
           class(out) <- classes[x] ; out }), NAMES)
       temp_metadata[["ColName"]] <- make.names(gsub("\\s", "", temp_metadata[["ColName"]]))
       temp_metadata
-  }
+}
 
-  read.dat <- function(dat, metadata_var, labels.included = "yes") {
+read.dat <- function(dat, metadata_var, labels.included = "yes") {
       read.fwf(dat, widths = metadata_var[["ColWidth"]], sep = "", col.names = metadata_var[["ColName"]])
-  }
+}
 
 
-GSS_metadata <- read.dct(here::here("inputs/data/gss-norc/GSS.dct"))
-GSS_ascii <- read.dat(here::here("inputs/data/gss-norc/GSS.dat"), GSS_metadata)
+GSS_metadata <- read.dct(here::here("inputs/data/raw/GSS.dct"))
+GSS_ascii <- read.dat(here::here("inputs/data/raw/GSS.dat"), GSS_metadata)
 attr(GSS_ascii, "col.label") <- GSS_metadata[["ColLabel"]]
 GSS <- GSS_ascii |>
   select(BALLOT, HAPPY, GIVHMLSS, DIRECTNS, YEAR)
 
 write_csv(
   x = GSS,
-  file = here::here("inputs/data/raw_data.csv"),
+  file = here::here("inputs/data/raw/GSS.csv"),
 )

### 2023-03-12 scripts/02-data_cleaning.r

@@ -1,0 +1,104 @@
+#### Preamble ####
+# Purpose: Import raw csv data, clean it, and generate new csv files
+# Author: Sebastian Rodriguez
+# Email: me@srod.ca
+# Date: 3 February 2023
+# GitHub: https://github.com/seb646/toronto-paramedic-responses
+
+#### Workspace set-up ####
+library(foreign)
+library(tidyverse)
+library(dplyr)
+library(here)
+library(readr)
+library(janitor)
+
+#### Fetch and clean happiness data (where happy) ####
+clean_happy_positive_data <- readr::read_csv(here::here("inputs/data/raw/GSS.csv"), show_col_types = FALSE) |>
+  # Change column naming format
+  clean_names() |> 
+  select(year, happy) |>
+  filter(happy >= 2) |> 
+  count(year, happy) |>
+  group_by(year) |>
+  summarise(people = sum(n))
+
+write_csv(
+  x = clean_happy_positive_data,
+  file = here::here("inputs/data/clean/happy_positive_data.csv"),
+)
+
+#### Fetch and clean directions data (where happy) ####
+clean_directions_positive_data <- readr::read_csv(here::here("inputs/data/raw/GSS.csv"), show_col_types = FALSE) |>
+  # Change column naming format
+  clean_names() |> 
+  select(year, directns) |>
+  filter(directns >= 1, directns <= 4) |> 
+  count(year, directns) |>
+  group_by(year) |>
+  summarise(people = sum(n))
+
+write_csv(
+  x = clean_directions_positive_data,
+  file = here::here("inputs/data/clean/directions_positive_data.csv"),
+)
+
+#### Fetch and homeless directions data (where happy) ####
+clean_homeless_positive_data <- readr::read_csv(here::here("inputs/data/raw/GSS.csv"), show_col_types = FALSE) |>
+  # Change column naming format
+  clean_names() |> 
+  select(year, givhmlss) |>
+  filter(givhmlss >= 1, givhmlss <= 4) |>
+  count(year, givhmlss) |>
+  group_by(year) |>
+  summarise(people = sum(n))
+
+write_csv(
+  x = clean_homeless_positive_data,
+  file = here::here("inputs/data/clean/homeless_positive_data.csv"),
+)
+
+#### Fetch and clean happiness data (where happy) ####
+clean_happy_negative_data <- readr::read_csv(here::here("inputs/data/raw/GSS.csv"), show_col_types = FALSE) |>
[…]

### 2023-03-12 outputs/paper/paper.qmd
Update attributions

@@
 ---
-title: "Does Happiness Lead to Altruism: A review of US Generial Social Survey data"
+title: "Does Happiness Lead to Altruism: A review of US General Social Survey data"
 format: pdf
@@
 #### Preamble ####
-# Title: The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?
-# Author: Sebastian Rodriguez
+# Title: Does Happiness Lead to Altruism: A review of US General Social Survey data
+# Author: Sebastian Rodriguez, Iz Leitch, Randall Ni
 # Email: me@srod.ca
 # Date: 3 February 2023
-# GitHub: https://github.com/seb646/toronto-paramedic-responses
+# GitHub: https://github.com/seb646/happiness-and-altruism
 
@@
 
-The General Social Survey (GSS) is a project of the independent research organization NORC at the University of Chicago, with principal funding from the National Science Foundation.
-
 **Resources Acknowledgment**

### 2023-03-12 scripts/03-data_testing.qmd
Update attributions

@@
 ---
-title: "Testing Data from the Toronto Paramedic Service"
+title: "Testing Data from the GSS"
 format: html
-author: "Sebastian Rodriguez"
+author: "Sebastian Rodriguez, Iz Leitch, Randall Niz"
 ---
@@
 #### Preamble ####
-# Purpose: Testing Data from the Toronto Paramedic Service
+# Purpose: Testing Data from the GSS
 # Prerequisites: Need to have downloaded the data using [inputs/data/data_import.qmd]
-# Author: Sebastian Rodriguez
+# Author: Sebastian Rodriguez, Iz Leitch, Randall Ni
 # Email: me@srod.ca
 # Date: 3 February 2023
-# GitHub: https://github.com/seb646/toronto-paramedic-responses
+# GitHub: https://github.com/seb646/happiness-and-altruism
 

### 2023-03-12 scripts/04-data_simulation.qmd
Update attributions

@@
 ---
-title: "Simulating Data from the Toronto Paramedic Service"
+title: "Simulating Data from the GSS"
 format: html
-author: "Sebastian Rodriguez"
+author: "Sebastian Rodriguez, Iz Leitch, Randall Ni"
 ---
@@
 #### Preamble ####
-# Purpose: Read in data from the Toronto Paramedic Service
-# Prerequisites: Need to know where to get Toronto Paramedic Services data
-# Author: Sebastian Rodriguez
+# Purpose: Read in data from the GSS
+# Prerequisites: Need to know where to get GSS data
+# Author: Sebastian Rodriguez, Iz Leitch, Randall Ni
 # Email: me@srod.ca
 # Date: 3 February 2023
-# GitHub: https://github.com/seb646/toronto-paramedic-responses
+# GitHub: https://github.com/seb646/happiness-and-altruism
 

### 2023-03-12 scripts/00-install_dependencies.r

@@ -1,19 +1,11 @@
----
-title: "Install the dependencies required for this project"
-format: html
-author: "Sebastian Rodriguez"
----
-```{r}
 #### Preamble ####
 # Purpose: Install the dependencies required for this project
 # Prerequisites: You must also have the R language installed
-# Author: Sebastian Rodriguez
+# Author: Sebastian Rodriguez, Iz Leitch, Randall Ni
 # Email: me@srod.ca
 # Date: 3 February 2023
-# GitHub: https://github.com/seb646/toronto-paramedic-responses
-```
+# GitHub: https://github.com/seb646/happiness-and-altruism
 
-```{r}
 # Install required packages (you only need to do this once)
 install.packages("lubridate")
 install.packages("tidyverse")
@@ -27,4 +19,3 @@ install.packages("kableExtra")
 install.packages("here")
 install.packages("readr")
 install.packages("foreign")
-```

### 2023-03-12 scripts/01-data_convert.r

@@ -1,9 +1,9 @@
 #### Preamble ####
-# Purpose: Import data from the Toronto Paramedic Service via Open Data Toronto
-# Author: Sebastian Rodriguez
+# Purpose: Import data from the GSS
+# Author: Sebastian Rodriguez, Iz Leitch, Randall Ni
 # Email: me@srod.ca
 # Date: 3 February 2023
-# GitHub: https://github.com/seb646/toronto-paramedic-responses
+# GitHub: https://github.com/seb646/happiness-and-altruism
 
 #### Workspace set-up ####
 library(foreign)

### 2023-03-12 scripts/02-data_cleaning.r

@@ -1,9 +1,9 @@
 #### Preamble ####
 # Purpose: Import raw csv data, clean it, and generate new csv files
-# Author: Sebastian Rodriguez
+# Author: Sebastian Rodriguez, Iz Leitch, Randall Ni
 # Email: me@srod.ca
 # Date: 3 February 2023
-# GitHub: https://github.com/seb646/toronto-paramedic-responses
+# GitHub: https://github.com/seb646/happiness-and-altruism
 
 #### Workspace set-up ####
 library(foreign)

### 2023-03-15 outputs/paper/paper.qmd
Updated paper graphs

@@
 #| warning: false
-#| fig-cap: "Happiness vs Altruism"
+#| fig-cap: "Happiness vs Giving Directions to Strangers"
 #| label: fig-1
@@
   geom_line(data = directions_positive_data, aes(x = year, y = people, color = "Give Directions to Stranger")) +
+  scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
+  scale_color_manual(values = colors)
+```
+
+```{r}
+#| messages: false
+#| echo: false
+#| warning: false
+#| fig-cap: "Happiness vs Giving Money to Homeless"
+#| label: fig-2
+
+colors <- c(
+  "Happiness" = "steelblue",
+  "Give Directions to Stranger" = "darkred",
+  "Give Money to Homeless" = "darkgreen"
+)
+
+ggplot() +
+  # Define axis labels
+  labs(
+    y = "People",
+    x = "Year",
+    color = "Legend"
+  ) +
+
+  # Set the theme
+  theme_minimal() +
+
+  geom_line(data = happy_positive_data, aes(x = year, y = people, color = "Happiness")) +
   geom_line(data = homeless_positive_data, aes(x = year, y = people, color = "Give Money to Homeless")) +
@@
 #| warning: false
-#| fig-cap: "Unhappiness vs. No Altruism"
-#| label: fig-2
+#| fig-cap: "Unhappiness vs. Not Giving Directions to Strangers"
+#| label: fig-3
 
@@
   geom_line(data = directions_negative_data, aes(x = year, y = people, color = "Give Directions to Stranger")) +
+  scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
+  scale_color_manual(values = colors)
+```
+
+```{r}
+#| messages: false
+#| echo: false
+#| warning: false
+#| fig-cap: "Unhappiness vs. Not Money to Homeless"
+#| label: fig-4
+
+colors <- c(
+  "Happiness" = "steelblue",
+  "Give Directions to Stranger" = "darkred",
+  "Give Money to Homeless" = "darkgreen"
+)
+
+ggplot() +
+  # Define axis labels
+  labs(
+    y = "People",
+    x = "Year",
+    color = "Legend"
+  ) +
+
+  # Set the theme
+  theme_minimal() +
+
+  geom_line(data = happy_negative_data, aes(x = year, y = people, color = "Happiness")) +
   geom_line(data = homeless_negative_data, aes(x = year, y = people, color = "Give Money to Homeless")) +

### 2023-03-16 outputs/paper/paper.qmd
Updates to paper content and structure

--- 2023-03-12 outputs/paper/paper.qmd
@@
 ---
-title: "Does Happiness Lead to Altruism: A review of US General Social Survey data"
+title: "Happiness Does Not Cause Altruism: A review of US General Social Survey data"
 format: pdf
-author: "Sebastian Rodriguez, Iz Leitch, Randall Ni"
-abstract: "Something about happiness and altruism" 
+author: "Sebastian Rodriguez, Randall Ni, Iz Leitch"
+abstract: "The relationship between happiness and altruistic behaviors has been the subject of much research, with some studies suggesting a positive correlation and others finding no significant relationship.
+This paper examines the existing literature on the topic in addition to the survey results provided by NORC’s General Social Survey (GSS) of United States residents and analyzes the data to draw conclusions.
+Based on the available evidence, it appears that there is little to no relationship between happiness and altruistic behaviors, suggesting that individuals may engage in altruistic acts for reasons other than personal happiness."
 date: today
@@
 # Title: Does Happiness Lead to Altruism: A review of US General Social Survey data
-# Author: Sebastian Rodriguez, Iz Leitch, Randall Ni
+# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
 # Email: me@srod.ca
-# Date: 3 February 2023
+# Date: 16 March 2023
 # GitHub: https://github.com/seb646/happiness-and-altruism
@@
 
+There is an implied correlation between emotions of happiness and altruistic behaviors.
+Happiness is a question of an individual's well-being, determined by emotional stability, feelings of satisfaction, and countless immeasurable qualitative factors.
+Happiness is also subjective, varying by an individual's definition of the term in addition to other factors like time of day, a given situation, and more.
+Altruism is a far more measured metric: while types of altruism differ, altruistic actions themselves are defined and can be quantified.
+Generally, a leading presumption is that feelings of happiness give a person the security and ability to take altruistic actions.
+
+Researchers have long sought to find a relationship between happiness and altruism, leading to mixed results.
+Some scholars argue that a "strong correlation exists" between the two, with caveats like "so long as they are not overwhelmed by helping tasks" [@cite_study_1].
+One investigation, focusing on university-aged people, found a "significant relationship" between happiness and altruism, going so far as to say that happiness "has a crucial role in creating altruism" [@cite_study_2].
+Other scholars argue that happiness has no bearing on altruistic behavior, and instead other feelings, like elevation, contribute to altruistic acts [@cite_study_3].
+These differing opinions are echoed throughout the research community, with evidence to support both hypotheses.
+
+Despite general feelings that happiness leads to altruism, there is an important question of whether there is actual causation or a mere correlation between the two social measures.
+This paper investigates levels of happiness against two specific altruistic acts: donating food or money to the homeless and giving directions to strangers.
+Specifically, this paper aims to interrogate the relationship between happiness and altruism using data from the NORC's General Social Survey (GSS) of United States residents.
+
+The findings revealed little evidence to support any relationship between happiness and altruism.
+Using survey results from 2002, 2004, 2012, and 2014, the data shows that there is no definitive link between feelings of happiness and altruistic acts.
+The survey data showed that, when feelings of happiness increased, it did not necessarily lead to the two altruistic acts studied in this paper.
+Although no link was found, it is worth noting that the question of happiness and altruism was not the purpose of these surveys.
+Both happiness and altruism were measured in the general survey, but it did not seek to study their relationship nor did it provide supplementary questions to aid in such a research endeavor.
+Further studies should explicitly seek to question the public's feelings of happiness against their altruistic behavior.\newpage
+
+## General Social Survey Data
+
+The primary source for data in this study was the United States General Social Survey (GSS), conducted by the University of Chicago's NORC with funding from the National Science Foundation.
+The GSS has been in operation since 1972, with consistent sampling and questioning that allows researchers to view and predict trends in data.
+The survey includes recurring and topical modules, with each participant receiving a subset of these modules.
+
+While the GSS consists of hundreds of questions covering a plethora of topics, this study examined three specific questions:
+
+1.  "Taken all together, how would you say things are these days---would you say that you are very happy, pretty happy, or not too happy?"
+2.  "During the past 12 months, how often have you... given directions to a stranger?"
+3.  "During the past 12 months, how often have you... given food or money to a homeless person?"
+
+Question 1 allows participants to respond with "Very happy", "Pretty Happy", "Not too happy", and a variety of abstentions.
+Questions 2 and 3 sought to examine the frequency of altruistic action, providing responses ranging from "More than once a week" to "Not at all in the past year", with a variety of abstention options.
+
+The survey's question about happiness was the only variable that tracked general happiness with the most complete set of data.
+Other options included happiness of marriage, happiness in relation to "today" (the day the survey was taken), and a sparsely collected "how happy would you say you are, on the whole." The other datasets either did not relate to this study's research question or lacked the data to provide meaningful insights.
+
+Questions about altruism had a significantly greater range when compared to questions about happiness.
+The survey's questions on altruism ranged from social actions to financial donations.
+The two questions selected for study, giving directions to strangers and giving food or money to the homeless, offered the greatest range of representative responses---it combined both social and financial altruism.
 \newpage
 
-## General Social Survey Data
+### Happiness and Donating to the Homeless
+
+Assisting those in need is the most fundamental definition of altruism, and donating to the homeless is perhaps the best representation of altruism.
+Donating to the homeless---whether in the form of money or food---requires empathy and true care for the well-being of others.
+While it can be considered a form of financial altruism, donating to the homeless goes far beyond other acts of financial altruism, like donating to charities.
+Donating to the homeless requires direct interactions with a homeless person.
+These interactions can be more difficult, there are time and physical restrictions that may prevent this form of altruism.
+However, this form of altruism represents the most direct and personal altruistic action a person can take.
+
+The first graph (@fig-1) charts the number of people who identified as "Very happy" or "Pretty happy" versus the number of people who donated food or money to a homeless person at least or more than 2 times per year.
+This frequency of altruistic action was selected as a baseline due to the possible restrictions that come with this action.
+For example, as explored earlier, financial and other barriers may prevent this type of altruism.
+The graph charts these two variables against the years 2002, 2004, 2012, and 2014.
 
@@
 #| warning: false
-#| fig-cap: "Happiness vs Altruism"
+#| fig-cap: "Happiness vs Giving Food or Money to the Homeless"
 #| label: fig-1
@@
 colors <- c(
-  "Happiness" = "steelblue",
-  "Give Directions to Stranger" = "darkred",
-  "Give Money to Homeless" = "darkgreen"
+  'Feeling "pretty happy" or "very happy"' = "steelblue",
+  "Giving money or food to the homeless\n more than 2 times a year" = "darkred"
 )
@@
   labs(
-    y = "People",
+    y = "Number of People",
     x = "Year",
@@
 
-  geom_line(data = happy_positive_data, aes(x = year, y = people, color = "Happiness")) +
-  geom_line(data = directions_positive_data, aes(x = year, y = people, color = "Give Directions to Stranger")) +
-  geom_line(data = homeless_positive_data, aes(x = year, y = people, color = "Give Money to Homeless")) +
+  geom_line(data = happy_positive_data, aes(x = year, y = people, color = 'Feeling "pretty happy" or "very happy"')) +
+  geom_line(data = homeless_positive_data, aes(x = year, y = people, color = "Giving money or food to the homeless\n more than 2 times a year")) +
   scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
@@
 ```
+
+\newpage
+
+The second graph (@fig-2) charts the number of people who identified as "Not too happy" against the number of people who did not donate food or money to a homeless person or only did so once a year.
+This "negative" view of unhappiness and non-altruistic behavior contrasts the previous graph and was intended to provide insight as to whether feeling not happy impacted a lack of altruistic behavior.
 
@@
 #| warning: false
-#| fig-cap: "Unhappiness vs. No Altruism"
+#| fig-cap: "Unhappiness vs. Not Giving Food or Money to the Homeless"
 #| label: fig-2
@@
 colors <- c(
-  "Happiness" = "steelblue",
-  "Give Directions to Stranger" = "darkred",
-  "Give Money to Homeless" = "darkgreen"
+  'Feeling "not to happy"' = "steelblue",
+  "Giving money or food to the homeless\n less than 2 times a year" = "darkred"
 )
@@
 
-  geom_line(data = happy_negative_data, aes(x = year, y = people, color = "Happiness")) +
-  geom_line(data = directions_negative_data, aes(x = year, y = people, color = "Give Directions to Stranger")) +
-  geom_line(data = homeless_negative_data, aes(x = year, y = people, color = "Give Money to Homeless")) +
+  geom_line(data = happy_negative_data, aes(x = year, y = people, color = 'Feeling "not to happy"')) +
+  geom_line(data = homeless_negative_data, aes(x = year, y = people, color =  "Giving money or food to the homeless\n less than 2 times a year")) +
   scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
@@
 ```
+
+Charting both happiness versus altruism and unhappiness versus non-altruistic tendencies revealed that there was no significant correlation between the two variables.
+Between 2002 and 2004, general happiness increased while this altruistic action decreased.
+However, between 2012 and 2014, both happiness and donations to the homeless increased, but they did not do so at the same or even a similar rate so it is insufficient in providing meaningful insights.
+
+\newpage
+
+### Happiness and Giving Directions to Strangers
+
+Altruism comes in a diverse range of actions---it represents more than financial or physical donations.
+A significant portion of the altruistic questions of the GSS relates to social altruism, questioning whether people donate their own time and energy to improve the well-being of others.
+Stopping to help a stranger comes with drawbacks: it is potentially dangerous, it consumes time, and it disrupts a person's routine.
+It also represents an arguably less important act when compared to helping the homeless---the stakes are lowered, and it is less necessary to society.
+Taking the time to stop and help a stranger when they are not in dire need requires an elevated altruistic sense that transcends mere obligation.
+
+The first graph (@fig-3) charts the number of people who identified as "Very happy" or "Pretty happy" versus the number of people who gave directions to a stranger at least or more than 2 times per year.
+Like donating to the homeless, this baseline was selected because of the possible restrictions that come with the action.
+Not everyone is in a position to give directions to a stranger, either because they live in a low-traffic, low-tourist area or because of other time restrictions.
+The graph charts these two variables in the years 2002, 2004, 2012, and 2014.
+
+```{r}
+#| messages: false
+#| echo: false
+#| warning: false
+#| fig-cap: "Happiness vs Giving Directions to Strangers"
+#| label: fig-3
+
+colors <- c(
+  'Feeling "pretty happy" or "very happy"' = "steelblue",
+  "Giving directions to a stranger\n more than 2 times a year" = "darkred"
+)
+
+ggplot() +
+  # Define axis labels
+  labs(
+    y = "Number of People",
+    x = "Year",
+    color = "Legend"
+  ) +
+
+  # Set the theme
+  theme_minimal() +
+
+  geom_line(data = happy_positive_data, aes(x = year, y = people, color = 'Feeling "pretty happy" or "very happy"')) +
+  geom_line(data = directions_positive_data, aes(x = year, y = people, color = "Giving directions to a stranger\n more than 2 times a year")) +
+  scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
+  scale_color_manual(values = colors)
+```
+
+\newpage
+
+The second graph (@fig-4) charts the number of people who said they were "Not too happy" against the number of people who did not give directions to strangers or only did so once a year.
+Highlighting this "negative" frame of the question seeks to understand if there is a correlation between a lack of happiness and non-altruistic behavior.
+
+```{r}
+#| messages: false
+#| echo: false
+#| warning: false
+#| fig-cap: "Unhappiness vs. Not Giving Directions to Strangers"
+#| label: fig-4
+
+colors <- c(
+  'Feeling "not too happy"' = "steelblue",
+  "Giving directions to a stranger\n less than 2 times a year" = "darkred"
+)
+
+ggplot() +
+  # Define axis labels
+  labs(
+    y = "Number of People",
+    x = "Year",
+    color = "Legend"
+  ) +
+
+  # Set the theme
+  theme_minimal() +
+
+  geom_line(data = happy_negative_data, aes(x = year, y = people, color = 'Feeling "not too happy"')) +
+  geom_line(data = directions_negative_data, aes(x = year, y = people, color = "Giving directions to a stranger\n less than 2 times a year")) +
+  scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
+  scale_color_manual(values = colors)
+```
+
+Charting happiness against the altruistic tendency to give directions to a stranger revealed that there was no significant correlation between happiness and altruism.
+In fact, the data shows that the opposite was true in this specific case.
+Between 2002 and 2004, feelings of happiness decreased while the act of giving directions to a stranger increased.
+Between 2012 and 2014, the number of people who gave directions to a stranger decreased steadily while there was a statistically significant number of people who reported feeling "Pretty happy" or "Very happy." When examining the negatives of both scenarios (unhappiness and not giving directions to a stranger), both the number of people who felt unhappy and the number of people who did not give directions significantly increased.
+\newpage
+
 ## Discussion
+
+Applying the U.S.
+General Social Survey (GSS) to this study came with significant limitations.
+The most notable limitation was that the survey was not intended to reveal a connection between happiness and altruism.
+The general nature of the survey meant that the questions were ambiguous and did not seek further detail from the participant.
+For example, while it asked the participant to label their happiness, the cause of their happiness---or unhappiness---was not clear.
+Further, altruistic questions were only asked in certain years: 2002, 2004, 2012, and 2014.
+This limited data set did not give the breadth of data that this study desired, although it was enough to make a preliminary determination on the relationship between happiness and altruism.
+
+The GSS questions also came with certain biases that may have skewed results.
+For example, in relation to altruism and giving directions to a stranger, it is important to ask: how often is the participant given the opportunity to perform this act?
+While many people in the United States live in a city or in high-traffic, high-tourist areas, a great number of people live in rural areas where they are not given such an opportunity.
+To the same extent, the question of opportunity is similar for donating food or money to the homeless.
+It is necessary to understand why a person did not donate to the homeless.
+Did they not have the opportunity because they do not live in an area that has a significant number of homeless people?
+Did they not have the financial means to donate at that time?
+Did they have the desire to donate, but did the act of traveling to an ATM or restaurant outweigh their altruistic urge?
+This context would reveal significant insight into the situations surrounding these altruistic actions and help determine whether the altruistic nature existed and the person simply lacked the means to follow through.
+
+Another important consideration is whether the GSS is an effective tool in answering this question.
+It was an effective tool---to an extent.
+The GSS provided a surface-level baseline to measure both happiness and altruistic actions.
+It collected self-reported levels of happiness and self-reported acts of altruism.
+This was enough to determine that there was rarely a correlation and no causation between happiness and altruism in the United States at the time of these surveys.
+However, it creates questions about altruistic intent and intrinsic definitions of happiness, while simultaneously providing a platform for further research.
+
+While this study suggests a lack of a relationship between happiness and altruism, the limitations that come from the General Social Survey (GSS) questions make it difficult to investigate the question further.
 
@@
 The primary tool used to analyse data in this paper is R, an open-source statistical programming language [@cite_tool_r].
-The paper also uses a number of R packages, including: dplyr [@cite_tool_dplyr], ggplot2 [@cite_tool_ggplot2], here [@cite_tool_here], janitor [@cite_tool_janitor], kableExtra [@cite_tool_kableextra], knitr [@cite_tool_knitr], lubridate [@cite_tool_lubridate], opendatatoronto [@cite_tool_opendatatoronto], readr [@cite_tool_readr], RColorBrewer [@cite_tool_rcolorbrewer], scales [@cite_tool_scales], and tidyverse [@cite_tool_tidyverse].
+The paper also uses a number of R packages, including: dplyr [@cite_tool_dplyr], foreign [@cite_tool_foreign], ggplot2 [@cite_tool_ggplot2], here [@cite_tool_here], janitor [@cite_tool_janitor], kableExtra [@cite_tool_kableextra], knitr [@cite_tool_knitr], lubridate [@cite_tool_lubridate], readr [@cite_tool_readr], RColorBrewer [@cite_tool_rcolorbrewer], scales [@cite_tool_scales], and tidyverse [@cite_tool_tidyverse].
 \newpage
+
+## Appendix
+
+### Appendix A: Supplementary Survey
+Researchers have long sought to find a connection between happiness and altruistic actions.
+Some studies show that happiness may lead to greater levels of altruism, while others fail to find a relationship between the emotion and subsequent action.
+With homelessness on the rise and a growing separation between socio-economic classes, it is important to discover the leading causes of altruism.
+
+By participating in this survey, you consent to us collecting your basic information and understand that we will be using your responses in our study.
+The survey is anonymous, we will not collect or share any personal identifiable information.
+The survey is also voluntary, and you may withdraw at any time.
+
+If you have any questions or concerns, feel free to contact Sebastian Rodriguez (s [dot] rodriguez [at] mail [dot] utoronto [dot] ca).
+
+1.  On a scale of 1-5, how happy do you feel right now?
+    a. Very happy
+    b. Pretty happy
+    c. Happy
+    d. Not too happy
+    e. Unhappy
+2.  How often have you given directions to a stranger in the past year?
+    a. More than once a week
+    b. Once a week
+    c. Once a month 
+    d. At least 2 or 3 times in the past year
+    e. Once in the past year
+    f. Not at all in the past year
+3.  How often are you given the opportunity to give directions to a stranger?
+    a. More than once a week
+    b. Once a week
+    c. Once a month 
+    d. At least 2 or 3 times in the past year
+    e. Once in the past year
+    f. Not at all in the past year
+4.  If did not give directions to a stranger, did you want to?
+    a. Yes
+    b. It depends on the day
+    c. It depends on the time I have in the moment
+    d. No
+5.  If you had the desire but did not give directions to a stranger, what prevented you from doing so?
+    a. I do not live in an area with a lot of strangers asking for directions
+    b. I did not have the time to help
+    c. I did not know enough about the area to be helpful
+    d. Another factor beyond my control prevented me from helping
+6.  How often have you donated food or money to a homeless person in the past year?
+    a. More than once a week
+    b. Once a week
+    c. Once a month 
+    d. At least 2 or 3 times in the past year
+    e. Once in the past year
+    f. Not at all in the past year
+7.  How often are you given the opportunity to donate to a homeless person?
+    a. More than once a week
+    b. Once a week
+    c. Once a month 
+    d. At least 2 or 3 times in the past year
+    e. Once in the past year
+    f. Not at all in the past year
+8.  If did not donate to a homeless person, did you want to?
+    a. Yes
+    b. It depends on the day
+    c. It depends on the money and resources at my disposal
+    d. No
+9.  If you had the desire but did not donate to a homeless person, what prevented you from doing so?
+    a. I do not live in an area with a lot of homeless people
+    b. I did not have the time to help
+    c. I did not have the money or resources to help
+    d. Another factor beyond my control prevented me from helping
+
+Thank you for participating in this survey!
+Your responses will greatly aid our effort to understand altruism and whether happiness has any impact on the behavior.
+If at any time you wish to withdraw from this survey, please contact Sebastian Rodriguez (s [dot] rodriguez [at] mail [dot] utoronto [dot] ca).
 

### 2023-03-16 outputs/paper/paper.qmd
Major updates
* Updated comments and metadata throughout project
* Updated readme files throughout project
* Added tests for the data
* Added data simulation

@@
 #### Preamble ####
-# Title: Does Happiness Lead to Altruism: A review of US General Social Survey data
+# Title: Happiness Does Not Cause Altruism: A review of US General Social Survey data
 # Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
@@
 
-# Fetch the the clean response data
+# Fetch the the clean happiness positive data
 happy_positive_data <- readr::read_csv(here::here("inputs/data/clean/happy_positive_data.csv"), show_col_types = FALSE)
 
-# Fetch clean performance data
+# Fetch clean directions positive data
 directions_positive_data <- readr::read_csv(here::here("inputs/data/clean/directions_positive_data.csv"), show_col_types = FALSE)
 
-# Fetch the clean incident response
+# Fetch the clean homeless positive data
 homeless_positive_data <- readr::read_csv(here::here("inputs/data/clean/homeless_positive_data.csv"), show_col_types = FALSE)
 
-# Fetch the the clean response data
+# Fetch the the clean happy negative data
 happy_negative_data <- readr::read_csv(here::here("inputs/data/clean/happy_negative_data.csv"), show_col_types = FALSE)
 
-# Fetch clean performance data
+# Fetch clean directions negative data
 directions_negative_data <- readr::read_csv(here::here("inputs/data/clean/directions_negative_data.csv"), show_col_types = FALSE)
 
-# Fetch the clean incident response
+# Fetch the clean homeless negative data
 homeless_negative_data <- readr::read_csv(here::here("inputs/data/clean/homeless_negative_data.csv"), show_col_types = FALSE)
@@
 
+# Set the colors for the lines and labels for the legend
 colors <- c(
@@
 
+  # Import the happy positive data for the line
   geom_line(data = happy_positive_data, aes(x = year, y = people, color = 'Feeling "pretty happy" or "very happy"')) +
+
+  # Import the homeless positive data for the line
   geom_line(data = homeless_positive_data, aes(x = year, y = people, color = "Giving money or food to the homeless\n more than 2 times a year")) +
+
+  # Set scales to only be years where data is provided
   scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
+
+  # Set colors for the graph
   scale_color_manual(values = colors)
@@
 
+  # Import the happy negative data for the line
   geom_line(data = happy_negative_data, aes(x = year, y = people, color = 'Feeling "not to happy"')) +
+
+  # Import the homeless negative data for the line
   geom_line(data = homeless_negative_data, aes(x = year, y = people, color =  "Giving money or food to the homeless\n less than 2 times a year")) +
+
+  # Set scales to only be years where data is provided
   scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
+
+  # Set colors for the graph
   scale_color_manual(values = colors)
@@
 
+  # Import the happy positive data for the line
   geom_line(data = happy_positive_data, aes(x = year, y = people, color = 'Feeling "pretty happy" or "very happy"')) +
+
+  # Import the directions positive data for the line
   geom_line(data = directions_positive_data, aes(x = year, y = people, color = "Giving directions to a stranger\n more than 2 times a year")) +
+
+  # Set scales to only be years where data is provided
   scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
+
+  # Set colors for the graph
   scale_color_manual(values = colors)
@@
 
+  # Import the happy negative data for the line
   geom_line(data = happy_negative_data, aes(x = year, y = people, color = 'Feeling "not too happy"')) +
+
+  # Import the directions negative data for the line
   geom_line(data = directions_negative_data, aes(x = year, y = people, color = "Giving directions to a stranger\n less than 2 times a year")) +
+
+  # Set scales to only be years where data is provided
   scale_x_continuous(breaks = c(2002, 2004, 2012, 2014)) +
+
+  # Set colors for the graph
   scale_color_manual(values = colors)

### 2023-03-16 scripts/00-install_dependencies.r

@@ -1,9 +1,9 @@
 #### Preamble ####
 # Purpose: Install the dependencies required for this project
 # Prerequisites: You must also have the R language installed
-# Author: Sebastian Rodriguez, Iz Leitch, Randall Ni
+# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
 # Email: me@srod.ca
-# Date: 3 February 2023
+# Date: 16 March 2023
 # GitHub: https://github.com/seb646/happiness-and-altruism
 
 # Install required packages (you only need to do this once)

### 2023-03-16 scripts/01-data_convert.r

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Import data from the GSS
-# Author: Sebastian Rodriguez, Iz Leitch, Randall Ni
+# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
 # Email: me@srod.ca
-# Date: 3 February 2023
+# Date: 16 March 2023
 # GitHub: https://github.com/seb646/happiness-and-altruism
 
 #### Workspace set-up ####

### 2023-03-16 scripts/02-data_cleaning.r

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Import raw csv data, clean it, and generate new csv files
-# Author: Sebastian Rodriguez, Iz Leitch, Randall Ni
+# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
 # Email: me@srod.ca
-# Date: 3 February 2023
+# Date: 16 March 2023
 # GitHub: https://github.com/seb646/happiness-and-altruism
 
 #### Workspace set-up ####

### 2023-03-16 scripts/03-data_testing.r

@@ -1,0 +1,114 @@
+#### Preamble ####
+# Purpose: Testing Data from the GSS
+# Prerequisites: Need to have downloaded and cleaned the data
+# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
+# Email: me@srod.ca
+# Date: 16 March 2023
+# GitHub: https://github.com/seb646/happiness-and-altruism
+
+#### Workspace set-up ####
+library(tidyverse)
+library(janitor)
+library(here)
+library(readr)
+
+#### Test the clean happiness positive data #### 
+# Fetch the the clean happiness positive data
+happy_positive_data <- readr::read_csv(here::here("inputs/data/clean/happy_positive_data.csv"), show_col_types = FALSE)
+
+# Test that the number of people in each column is less than or equal to 6,000
+happy_positive_data$people |> max() <= 6000
+
+# Test that the number of people in each column is greater than or equal to 1
+happy_positive_data$people |> min() >= 1
+
+# Test that the year is less than or equal to 2014
+happy_positive_data$year |> max() <= 2014
+
+# Test that the year is greater than or equal to 2002
+happy_positive_data$year |> min() >= 2002
+
+
+#### Test the clean directions positive data #### 
+# Fetch clean directions positive data
+directions_positive_data <- readr::read_csv(here::here("inputs/data/clean/directions_positive_data.csv"), show_col_types = FALSE)
+
+# Test that the number of people in each column is less than or equal to 6,000
+directions_positive_data$people |> max() <= 6000
+
+# Test that the number of people in each column is greater than or equal to 1
+directions_positive_data$people |> min() >= 1
+
+# Test that the year is less than or equal to 2014
+directions_positive_data$year |> max() <= 2014
+
+# Test that the year is greater than or equal to 2002
+directions_positive_data$year |> min() >= 2002
+
+
+#### Test the clean homeless positive data #### 
+# Fetch the clean homeless positive data
+homeless_positive_data <- readr::read_csv(here::here("inputs/data/clean/homeless_positive_data.csv"), show_col_types = FALSE)
+
+# Test that the number of people in each column is less than or equal to 6,000
+homeless_positive_data$people |> max() <= 6000
+
+# Test that the number of people in each column is greater than or equal to 1
+homeless_positive_data$people |> min() >= 1
+
+# Test that the year is less than or equal to 2014
+homeless_positive_data$year |> max() <= 2014
+
+# Test that the year is greater than or equal to 2002
[…]

### 2023-03-16 scripts/04-data_simulation.r

@@ -1,0 +1,53 @@
+### Preamble ####
+# Purpose: Read in data from the GSS
+# Prerequisites: Need to know where to get GSS data
+# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
+# Email: me@srod.ca
+# Date: 16 March 2023
+# GitHub: https://github.com/seb646/happiness-and-altruism
+
+#### Workspace set-up ####
+library(tidyverse) # A collection of data-related packages
+library(janitor) # Helps clean datasets
+set.seed(124) # Set the seed for consistent randomness
+
+#### Simulate data based on three datasets from GSS ####
+# General Happiness: https://gssdataexplorer.norc.org/variables/434/vshow
+# Has Given Directions to a Stranger: https://gssdataexplorer.norc.org/variables/2886/vshow
+# Has Given Money or Food to a Homeless Person: https://gssdataexplorer.norc.org/variables/2878/vshow
+
+simulated_raw_data <-
+  tibble(
+    # Configure for ballots numbered 1-3
+    "BALLOT" = sample(
+      x = c(1:3),
+      size = 10000,
+      replace = TRUE
+    ),
+    # Configure for happiness responses -100, -99, -98, -97, 1, 2, 3
+    "HAPPY" = sample(
+      x = c(-100, -99, -98, -97, 1, 2, 3),
+      size = 10000,
+      replace = TRUE
+    ),
+    # Configure for homeless giving responses -100, -99, -98, -97, 1, 2, 3, 4, 5, 6
+    "GIVHMLSS" = sample(
+      x = c(-100, -99, -98, -97, 1, 2, 3, 4, 5, 6),
+      size = 10000,
+      replace = TRUE
+    ),
+    # Configure for directions responses -100, -99, -98, -97, 1, 2, 3, 4, 5, 6
+    "DIRECTNS" = sample(
+      x = c(-100, -99, -98, -97, 1, 2, 3, 4, 5, 6),
+      size = 10000,
+      replace = TRUE
+    ),
+    # Configure for years 2002, 2004, 2012, 2014
+    "YEAR" = sample(
+      x = c(2002, 2004, 2012, 2014),
+      size = 10000,
+      replace = TRUE
+    )
+  )
+
+simulated_raw_data

### 2023-03-17 outputs/paper/paper.qmd
Updates to paper – survey

@@
 ### Appendix A: Supplementary Survey
+The supplementary survey attached to this project is available through Google Forms: https://forms.gle/Zr2jj93bXFaaPkPdA.
+
+#### Survey Introduction 
 Researchers have long sought to find a connection between happiness and altruistic actions.
@@
 
+#### Survey Questions
 1.  On a scale of 1-5, how happy do you feel right now?
@@
     f. Not at all in the past year
-4.  If did not give directions to a stranger, did you want to?
+4.  If you did not give directions to a stranger, did you want to?
     a. Yes
@@
     f. Not at all in the past year
-8.  If did not donate to a homeless person, did you want to?
+8.  If you did not donate to a homeless person, did you want to?
     a. Yes
@@
 
+#### Survey Conclusion
 Thank you for participating in this survey!

### 2023-03-17 outputs/paper/paper.qmd
Add raw data tables to paper, update attributions

@@
 format: pdf
-author: "Sebastian Rodriguez, Randall Ni, Iz Leitch"
+author: "Sebastian Rodriguez, Laura Lee-Chu, Iz Leitch"
 abstract: "The relationship between happiness and altruistic behaviors has been the subject of much research, with some studies suggesting a positive correlation and others finding no significant relationship.
@@
 # Title: Happiness Does Not Cause Altruism: A review of US General Social Survey data
-# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
+# Author: Sebastian Rodriguez, Laura Lee-Chu, Iz Leitch
 # Email: me@srod.ca
@@
 
-### Appendix A: Supplementary Survey
+### Appendix A: Raw Data
+
+#### Happiness
+```{r}
+#| messages: false
+#| echo: false
+#| warning: false
+#| fig-cap: "Raw GSS Data: General Happiness (by year)"
+#| label: fig-5
+#### Create a table summarizing happiness data from GSS ####
+summarized_happiness_data <- merge(x = happy_positive_data, y = happy_negative_data, by = "year", all.x = TRUE) |>
+   # Rename columns
+  rename("Year" = "year", '"Very" or "Pretty" happy' = "people.x", '"Not too happy"' = "people.y")
+
+summarized_happiness_data |>
+  # Table configuration
+  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
+
+  # Add headers for table data
+  add_header_above(
+    c(" ", "Number of People Feeling" = 2),
+    align = "c",
+    bold = TRUE
+  ) |>
+
+  # Style the table
+  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
+
+  # Make the table headers bold
+  row_spec(0, bold = TRUE) |>
+
+  # Increase width of first column
+  column_spec(1, width = "10em")
+```
+
+#### Donating to the Homeless
+```{r}
+#| messages: false
+#| echo: false
+#| warning: false
+#| fig-cap: "Raw GSS Data: Giving Food or Money to the Homeless (by year)"
+#| label: fig-6
+#### Create a table summarizing homeless donation data from GSS ####
+summarized_homeless_data <- merge(x = homeless_positive_data, y = homeless_negative_data, by = "year", all.x = TRUE) |>
+   # Rename columns
+  rename("Year" = "year", 'At least twice a year' = "people.x", 'Less than twice a year' = "people.y")
+
+summarized_homeless_data |>
+  # Table configuration
+  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
+
+  # Add headers for table data
+  add_header_above(
+    c(" ", "Number of People Donating to the Homeless" = 2),
+    align = "c",
+    bold = TRUE
+  ) |>
+
+  # Style the table
+  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
+
+  # Make the table headers bold
+  row_spec(0, bold = TRUE) |>
+
+  # Increase width of first column
+  column_spec(1, width = "10em")
+```
+
+#### Giving Directions to a Stanger
+```{r}
+#| messages: false
+#| echo: false
+#| warning: false
+#| fig-cap: "Raw GSS Data: Giving Directions to a Stanger (by year)"
+#| label: fig-7
+#### Create a table summarizing giving strangers directions data from GSS ####
+summarized_directions_data <- merge(x = directions_positive_data, y = directions_negative_data, by = "year", all.x = TRUE) |>
+   # Rename columns
+  rename("Year" = "year", 'At least twice a year' = "people.x", 'Less than twice a year' = "people.y")
+
+summarized_directions_data |>
+  # Table configuration
+  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
+
+  # Add headers for table data
+  add_header_above(
+    c(" ", "Number of People Giving Strangers Directions" = 2),
+    align = "c",
+    bold = TRUE
+  ) |>
+
+  # Style the table
+  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
+
+  # Make the table headers bold
+  row_spec(0, bold = TRUE) |>
+
+  # Increase width of first column
+  column_spec(1, width = "10em")
+```
+
+### Appendix B: Supplementary Survey
 The supplementary survey attached to this project is available through Google Forms: https://forms.gle/Zr2jj93bXFaaPkPdA.

### 2023-03-17 scripts/00-install_dependencies.r

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Install the dependencies required for this project
 # Prerequisites: You must also have the R language installed
-# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
+# Author: Sebastian Rodriguez, Laura Lee-Chu, Iz Leitch
 # Email: me@srod.ca
 # Date: 16 March 2023
 # GitHub: https://github.com/seb646/happiness-and-altruism

### 2023-03-17 scripts/01-data_convert.r

@@ -1,6 +1,6 @@
 #### Preamble ####
 # Purpose: Import data from the GSS
-# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
+# Author: Sebastian Rodriguez, Laura Lee-Chu, Iz Leitch
 # Email: me@srod.ca
 # Date: 16 March 2023
 # GitHub: https://github.com/seb646/happiness-and-altruism

### 2023-03-17 scripts/02-data_cleaning.r

@@ -1,6 +1,6 @@
 #### Preamble ####
 # Purpose: Import raw csv data, clean it, and generate new csv files
-# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
+# Author: Sebastian Rodriguez, Laura Lee-Chu, Iz Leitch
 # Email: me@srod.ca
 # Date: 16 March 2023
 # GitHub: https://github.com/seb646/happiness-and-altruism

### 2023-03-17 scripts/03-data_testing.r

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Testing Data from the GSS
 # Prerequisites: Need to have downloaded and cleaned the data
-# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
+# Author: Sebastian Rodriguez, Laura Lee-Chu, Iz Leitch
 # Email: me@srod.ca
 # Date: 16 March 2023
 # GitHub: https://github.com/seb646/happiness-and-altruism

### 2023-03-17 scripts/04-data_simulation.r

@@ -1,7 +1,7 @@
 ### Preamble ####
 # Purpose: Read in data from the GSS
 # Prerequisites: Need to know where to get GSS data
-# Author: Sebastian Rodriguez, Randall Ni, Iz Leitch
+# Author: Sebastian Rodriguez, Laura Lee-Chu, Iz Leitch
 # Email: me@srod.ca
 # Date: 16 March 2023
 # GitHub: https://github.com/seb646/happiness-and-altruism

### 2023-03-17 outputs/paper/paper.qmd
Added anchor links and structural fixes

@@
 bibliography: references.bib
-thanks: "Code and data supporting this analysis is available at: https://github.com/seb646/happiness-and-altruism"
+thanks: "Code and data supporting this analysis is available at: [https://github.com/seb646/happiness-and-altruism](https://github.com/seb646/happiness-and-altruism)"
 ---
@@
 
-We wish to acknowledge the land on which this data was collected and analysed.
+We wish to acknowledge the land on which this data was analysed.
 For thousands of years, it has been the traditional land of the Huron-Wendat, the Seneca, and the Mississaugas of the Credit.
@@
 
-[^1]: General Happiness data is available at: https://gssdataexplorer.norc.org/variables/434/vshow
+[^1]: General Happiness data is available at: [https://gssdataexplorer.norc.org/variables/434/vshow](https://gssdataexplorer.norc.org/variables/434/vshow)
 
-[^2]: Has Given Directions to a Stranger data is available at: https://gssdataexplorer.norc.org/variables/2886/vshow
+[^2]: Has Given Directions to a Stranger data is available at: [https://gssdataexplorer.norc.org/variables/2886/vshow](https://gssdataexplorer.norc.org/variables/2886/vshow)
 
-[^3]: Has Given Money or Food to a Homeless Person data is available at: https://gssdataexplorer.norc.org/variables/2878/vshow
+[^3]: Has Given Money or Food to a Homeless Person data is available at: [https://gssdataexplorer.norc.org/variables/2878/vshow](https://gssdataexplorer.norc.org/variables/2878/vshow)\
 
@@
 
-#### Happiness
+**Happiness**
+
 ```{r}
@@
 ```
+\
 
-#### Donating to the Homeless
+**Donating to the Homeless**
 ```{r}
@@
 ```
+\
 
-#### Giving Directions to a Stanger
+**Giving Directions to a Stranger**
+
 ```{r}
@@
 ```
+\newpage
 
 ### Appendix B: Supplementary Survey
-The supplementary survey attached to this project is available through Google Forms: https://forms.gle/Zr2jj93bXFaaPkPdA.
+The supplementary survey attached to this project is available through Google Forms: [https://forms.gle/Zr2jj93bXFaaPkPdA](https://forms.gle/Zr2jj93bXFaaPkPdA). \
 
-#### Survey Introduction 
+**Survey Introduction**
+
 Researchers have long sought to find a connection between happiness and altruistic actions.
@@
 
-If you have any questions or concerns, feel free to contact Sebastian Rodriguez (s [dot] rodriguez [at] mail [dot] utoronto [dot] ca).
+If you have any questions or concerns, feel free to contact Sebastian Rodriguez (s [dot] rodriguez [at] mail [dot] utoronto [dot] ca). \
 
-#### Survey Questions
+**Survey Questions**
+
 1.  On a scale of 1-5, how happy do you feel right now?
@@
     c. I did not have the money or resources to help
-    d. Another factor beyond my control prevented me from helping
+    d. Another factor beyond my control prevented me from helping\
 
-#### Survey Conclusion
+**Survey Conclusion**
+
 Thank you for participating in this survey!
@@
 
+\newpage
+
 ## References
